---
title: "Bleaching_legacy_effects_analyses"
author: "Marine Lechene"
date: "2026-06-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# 1. Effect of heat stress on survival and growth of healthy colonies

## 1.a. Survival and heat stress

We model Survival \~ Taxa \* Year to see whether survival was different under heat stress. We expect significantly higher mortality in year 2 (after bleaching).

```{r}
surv1 <- read_csv("C:/Users/jc815475/OneDrive - James Cook University/PhD data/3_Coral growth/Data/CSV/survival_transition1.csv", trim_ws=TRUE)
surv2 <- read_csv("C:/Users/jc815475/OneDrive - James Cook University/PhD data/3_Coral growth/Data/CSV/survival_transition2.csv", trim_ws=TRUE)
surv1 <- surv1 |> 
  mutate(Reef = factor(Reef),
         Site = factor(Site),
         Plot = factor(Plot),
         Colony = factor(Colony),
         Habitat = factor(Habitat),
         Depth = factor(Depth),
         Cross_shelf = factor(Cross_shelf),
         Taxa = factor(Taxa))

surv2 <- surv2 |> 
  mutate(Reef = factor(Reef),
         Site = factor(Site),
         Plot = factor(Plot),
         Colony = factor(Colony),
         Habitat = factor(Habitat),
         Depth = factor(Depth),
         Cross_shelf = factor(Cross_shelf),
         Taxa = factor(Taxa))

surv_1 <- surv1 |>
  filter(Taxa %in% c("Abra", "Acor", "Pdam", "Pver", "Shys"))
surv_2 <- surv2 |>
  filter(Taxa %in% c("Abra", "Acor", "Pdam", "Pver", "Shys"))

```

```{r}
# Prepare datasets to merge them back just to look at Survival ~ Year
surv_1$Year <- "2022"
surv_1$Bleaching_scores <- "1"
surv_2$Year <- "2023"

s1 <- surv_1 |> 
  rename(log_size = "log10_2021",
         Survival = "Survival_1")

s2 <- surv_2 |> 
  rename(log_size = "log10_2022",
         Survival = "Survival_2")

s1 <- s1 |> 
  dplyr::select(Reef, Site, Plot, Habitat, Depth, Cross_shelf, Taxa, Survival, Bleaching_scores, log_size, Year, Days_2022c)

s2 <- s2 |> 
  dplyr::select(Reef, Site, Plot, Habitat, Depth, Cross_shelf, Taxa, Survival, Bleaching_scores, log_size, Year, Days_2022c)

surv_data <- rbind(s1, s2)

surv_healthy <- surv_data |> 
  filter(Bleaching_scores == "1")

surv_healthy <- surv_healthy |> 
  mutate(Year = factor(Year),
         Plot = factor(Plot)) |> 
  droplevels()

surv_healthy |> 
  group_by(Taxa) |> 
  summarise(total_count = n()) #lowest sample size is Shys with 207 colonies
  
```

```{r}
#| label: model1.2.a. selection
#| warning: false
#| message: false
m0 <- glmmTMB(Survival ~ Taxa:Year + Days_2022c + (1|Reef/Site/Plot),
                 data = surv_healthy,
                 family = "binomial")
m1 <- glmmTMB(Survival ~ Taxa:Year + Days_2022c + (1|Reef/Site),
                 data = surv_healthy,
                 family = "binomial")
m2 <- glmmTMB(Survival ~ Taxa:Year + Days_2022c + (1|Reef/Plot),
                 data = surv_healthy,
                 family = "binomial")
m3 <- glmmTMB(Survival ~ Taxa:Year + Days_2022c + (1|Reef),
                 data = surv_healthy,
                 family = "binomial")
m4 <- glmmTMB(Survival ~ Taxa:Year + Days_2022c + (1|Site/Plot),
                 data = surv_healthy,
                 family = "binomial")
m5 <- glmmTMB(Survival ~ Taxa:Year + Days_2022c + (1|Site),
                 data = surv_healthy,
                 family = "binomial")
m6 <- glmmTMB(Survival ~ Taxa:Year + Days_2022c + (1|Plot),
                 data = surv_healthy,
                 family = "binomial")
```

```{r}
AIC(m0, m1, m2, m3, m4, m5, m6)
```

Best random effect structure:

```{r}
#| label: model surv1.2.a
#| message: false
#| warning: false
surv_heatstress <- glmmTMB(Survival ~ Taxa:Year + Days_2022c + (1|Reef/Plot),
                 data = surv_healthy,
                 family = "binomial")
```

```{r}
#| label: model surv1.2.a resids
#| message: false
#| warning: false
#| results: false
#| echo: false
surv_heatstress |> simulateResiduals(plot=T)
```

```{r}
#| label: model surv1.2.a preds
#| message: false
#| warning: false
#| echo: false
surv_heatstress_emm <- emmeans::emmeans(surv_heatstress, ~Year*Taxa, type = 'response')
surv_heatstress_emm <- as.data.frame(surv_heatstress_emm)
surv_heatstress_emm
```

## 1.b. Figure 2

```{r}
#| label: model surv1.2.a fig
#| echo: false
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 10

taxa_labels <- c("Abra" = "Acropora staghorn",
                 "Acor" = "Acropora corymbose",
                 "Pdam" = "Pocillopora damicornis",
                 "Pver" = "Pocillopora verrucosa",
                 "Shys" = "Seriatopora hystrix")

surv_heatstress_emm |> ggplot() +
  geom_point(aes(x = Year, y = prob, fill = Year, group = Taxa),
             size = 4, pch = 21)+
  geom_errorbar(aes(x = Year, ymin = asymp.LCL, ymax = asymp.UCL, 
                                   group = Taxa), width = 0.1, linewidth = 0.8) +
  scale_y_continuous(breaks = seq(0, 1.5, by = 0.25), limits = c(0.5, 1)) +
  scale_fill_manual(values = c("cadetblue", "darkorange"))+
  theme_bw() +
  facet_wrap(~ Taxa, labeller =  as_labeller(taxa_labels), ncol = 2)+
  labs(x = "", y = "Survival probability") +
  theme(
    axis.text.x = element_blank(),
    panel.grid.major.y = element_line(linetype = "dashed"),
    panel.grid.minor.y = element_blank()
  )
```

Survival probability tend to be lower in 2023 even for healthy colonies. Only *S. hystrix* shows significant differences in survival probability between 2022 and 2023. We could also say for A. staghorn as there was 0 mortality in 2021-2022.

## 1.c. Growth and heat stress

Here we modeled growth for only 'healthy' colonies (i.e., colonies that did not bleach in 2022) and only Davies and Palms so we could include them all in a single model:

logsize(t) \~ logsize(t previous) \* Year \* Taxa + (1\|Colony)

```{r}
#| label: healthymod dataprep
#| echo: false
#| message: false
#| warning: false

grow1 <- read_csv("C:/Users/jc815475/OneDrive - James Cook University/PhD data/3_Coral growth/Data/CSV/growth_transition1.csv", trim_ws=TRUE)
grow2 <- read_csv("C:/Users/jc815475/OneDrive - James Cook University/PhD data/3_Coral growth/Data/CSV/growth_transition2.csv", trim_ws=TRUE)
grow1 <- grow1 |> 
  mutate(Reef = factor(Reef),
         Site = factor(Site),
         Plot = factor(Plot),
         Colony = factor(Colony),
         Habitat = factor(Habitat),
         Depth = factor(Depth),
         Cross_shelf = factor(Cross_shelf),
         Taxa = factor(Taxa))

grow2 <- grow2 |> 
  mutate(Reef = factor(Reef),
         Site = factor(Site),
         Plot = factor(Plot),
         Colony = factor(Colony),
         Habitat = factor(Habitat),
         Depth = factor(Depth),
         Cross_shelf = factor(Cross_shelf),
         Taxa = factor(Taxa))

# Filter for taxa of interest
grow_1 <- grow1 |>
  filter(Taxa %in% c("Abra", "Acor", "Pdam", "Pver", "Shys"))
grow_2 <- grow2 |>
  filter(Taxa %in% c("Abra", "Acor", "Pdam", "Pver", "Shys"))
######

rawdata <- read_csv("C:/Users/jc815475/OneDrive - James Cook University/PhD data/3_Coral growth/Data/CSV/growth_surv_data.csv", trim_ws=TRUE)

dat1 <- rawdata |> 
  mutate(Taxa_2023 = factor(Class_2023),
         Taxa_2023 = fct_recode(Taxa_2023,
                                Atab = "Acyt",
                                Atab = "Ahya",
                                Abra = "Asta",
                                Adgt = "Agem",
                                Adgt = "Ahum",
                                Acor = "Amil",
                                Acor = "Aten",
                                Gspp = "Gedw",
                                Gspp = "Gret"),
         Taxa_2022 = factor(Class_2022),
         Taxa_2022 = fct_recode(Taxa_2022,
                                Atab = "Acyt",
                                Atab = "Ahya",
                                Abra = "Asta",
                                Adgt = "Agem",
                                Adgt = "Ahum",
                                Acor = "Amil",
                                Acor = "Aten",
                                Gspp = "Gedw",
                                Gspp = "Gret"),
         Taxa_2021 = factor(Class_2021),
         Taxa_2021 = fct_recode(Taxa_2021,
                                Atab = "Acyt",
                                Atab = "Ahya",
                                Abra = "Asta",
                                Adgt = "Agem",
                                Adgt = "Ahum",
                                Acor = "Amil",
                                Acor = "Aten",
                                Gspp = "Gedw",
                                Gspp = "Gret"))

### Subset data for each transition
healthy <- dat1 |> 
  filter(to_use_0 == "yes" & to_use_growth_1 == "yes" & to_use_growth_2 == "yes")

### Only retain taxa of interest
healthy <- healthy |> 
  filter(Taxa_2023 %in% c("Abra", "Acor", "Pdam", "Pver", "Shys")) |> 
  droplevels()

healthy <- healthy |> 
  filter(Taxa_2022 %in% c("Abra", "Acor", "Pdam", "Pver", "Shys")) |> 
  droplevels()

healthy <- healthy |> 
  filter(Taxa_2021 %in% c("Abra", "Acor", "Pdam", "Pver", "Shys")) |> 
  droplevels()

### only retain rows where Taxa match throughout the years
healthy <- healthy[healthy$Taxa_2021 == healthy$Taxa_2022 & healthy$Taxa_2022 == healthy$Taxa_2023, ]

### Check taxa levels for each year
levels(healthy$Taxa_2021)
levels(healthy$Taxa_2022)
levels(healthy$Taxa_2023)

### only retain colonies that stayed healthy in Year 2
healthy <- healthy |> 
  mutate(Bleaching_scores = factor(Bleaching_scores)) |> 
  filter(Bleaching_scores == "1")

## Setup hierarchical design
healthy <- healthy |> 
  mutate(Plot = factor(Plot_ID),
         Zone = factor(str_sub(Plot, 9,9)),
         Reef = factor(str_sub(Plot, 1,4)),
         Site = factor(str_sub(Plot, 1,9)), #unique combination of Reef/Site/Zone
         Site_type = factor(str_sub(Plot, 6, 7)),
         Colony_ID = row_number(),
         Colony = factor(paste(Plot, Colony_ID, sep = "_")),
         Habitat = factor(paste(Site_type, Zone, sep = "_")),
         Reef = fct_recode(Reef,
                           PA = "PAPE",
                           PA = "PAOR"),
         Depth = fct_recode(Zone,
                            Deep = "D",
                            Shallow = "S"),
         Area_cm_2021 = Area_2021*10000, #change area in m2 to cm2
         Area_cm_2022 = Area_2022*10000,
         Area_cm_2023 = Area_2023*10000,
         log_2021= log(Area_cm_2021),
         log_2022= log(Area_cm_2022),
         log_2023= log(Area_cm_2023),
         log10_2021=log10(Area_cm_2021),
         log10_2022 =log10(Area_cm_2022),
         log10_2023 =log10(Area_cm_2023),
         Date_2021 = factor(Date_2021),
         Date_2022 = factor(Date_2022),
         Taxa = factor(Taxa_2021),
         Cross_shelf = fct_recode(Reef,
                                  offshore = "OCCH",
                                  offshore = "OCDA",
                                  offshore = "OCLB",
                                  inshore = "PA"))


healthy <- healthy |> 
  mutate(
    Days = case_when(
      Date_2021 == "202101" ~ 365,
      Date_2021 == "202105" ~ 240, #add variable to account for different in days between surveys in 2021
      TRUE ~ NA_real_
    ),
    Days_c = Days - 365 #center on 365 days as the reference interval
  )

healthy <- healthy |> 
  dplyr::select(Reef, Site, Plot, Colony, Habitat, Depth, Cross_shelf, Taxa, Area_cm_2021, Area_cm_2022, Area_cm_2023, log_2021, log_2022, log_2023, log10_2021, log10_2022, log10_2023, Note_2021, Note_2022, Note_2023, Days_c)

growth_2022 <- healthy |> 
  transmute(
    Reef, Site, Plot, Colony, Habitat, Depth, Cross_shelf, Taxa,
    Year = factor("2022", levels = c("2022", "2023")),
    Area = Area_cm_2022,
    Area_prev = Area_cm_2021,
    log_size = log_2022,
    log_size_prev = log_2021,
    log10_size = log10_2022,
    log10_size_prev = log10_2021,
    Days_2022c = Days_c  
  )

growth_2023 <- healthy |> 
  transmute(
    Reef, Site, Plot, Colony, Habitat, Depth, Cross_shelf, Taxa,
    Year = factor("2023", levels = c("2022", "2023")),
    Area = Area_cm_2023,
    Area_prev = Area_cm_2022,
    log_size = log_2023,
    log_size_prev = log_2022,
    log10_size = log10_2023,
    log10_size_prev = log10_2022,
    Days_2022c = 0  
  )

healthy_model_data <- bind_rows(growth_2022, growth_2023) 

healthy_model_data |> 
  group_by(Taxa) |> 
  summarise(total_count = n()/2) #low number for Shys (4), Pdam (14)
```

Only 4 healthy colonies of S. hystrix to follow - caution when interpreting the results

```{r}
#| label: healthymod
#| message: false
#| warning: false
healthy_m1 <- glmmTMB(log_size ~ log_size_prev : Year : Taxa + Days_2022c + (1|Colony),
                     data = healthy_model_data)
healthy_m2 <- glmmTMB(log_size ~ log_size_prev : Year : Taxa + Days_2022c + (1|Plot/Colony),
                     data = healthy_model_data)
healthy_m3 <- glmmTMB(log_size ~ log_size_prev : Year : Taxa + Days_2022c + (1|Site/Plot/Colony),
                     data = healthy_model_data)
healthy_m4 <- glmmTMB(log_size ~ log_size_prev : Year : Taxa + Days_2022c + (1|Reef/Site/Plot/Colony),
                     data = healthy_model_data)
healthy_m5 <- glmmTMB(log_size ~ log_size_prev : Year : Taxa + Days_2022c + (1|Reef/Colony),
                     data = healthy_model_data)
healthy_m6 <- glmmTMB(log_size ~ log_size_prev : Year : Taxa + Days_2022c + (1|Site/Colony),
                     data = healthy_model_data)

```

```{r}
#| label: healthymod AIC
#| message: false
#| warning: false

AICc(healthy_m1, healthy_m2, healthy_m3, healthy_m4, healthy_m5, healthy_m6)
```

Best random effect structure is (1\|Site/Plot/Colony)

```{r}
#| label: healthymod summary
summary(healthy_m3)
```

```{r}
#| label: healthymod preds
#| echo: false
#| message: false
#| warning: false

healthy_m3.resid <- healthy_m3 |>  simulateResiduals(plot=T)

#Marginal effects
preds <- ggpredict(healthy_m3, terms = c("log_size_prev [all]", "Year", "Taxa"))
preds <- preds |> rename(Taxa = facet)
```

```{r}
#| label: healthymod emmeans
#| echo: false
#| message: false
#| warning: false
emm <- emmeans(
  healthy_m3,
  ~ Year | Taxa,
  at = list(log_size_prev = mean(healthy_model_data$log_size_prev)),
  type = "response"  # or type = "link" for log scale
)

pairwise_contrasts <- contrast(emm, method = "pairwise", adjust = "none") |>
  summary() |>
  mutate(Significance = ifelse(p.value < 0.05, "*", ""))

knitr::kable(pairwise_contrasts, digits = 5,
             caption = "Pairwise contrasts of Year within each coral taxon") |>
  kableExtra::kable_styling("striped", full_width = FALSE) |>
  kableExtra::row_spec(0, bold = TRUE)

```

All taxa except Pdam show significant difference in growth before and after heatstress

## 1.d. Figure 3

```{r}
#| label: healthymodfig
#| echo: false
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 10

preds$diff <- preds$predicted - preds$x
zero_crossings <- preds |>
  group_by(Taxa, group) |>
  arrange(abs(diff)) |>  # Sort by the smallest absolute difference
  slice(1)                # Take the closest point to the no-growth line


# Compute the range of 'log' for each Taxa
log_ranges <- healthy_model_data |>
  group_by(Taxa) |>
  summarize(min_log = min(log_size, na.rm = TRUE),
            max_log = max(log_size, na.rm = TRUE))

ggplot() +
  # Add shaded rectangles outside the real size range for each Taxa
  geom_rect(data = log_ranges, 
            aes(xmin = -Inf, xmax = min_log, ymin = -Inf, ymax = Inf), 
            fill = "grey", alpha = 0.2) +  # Left side greyed out
  geom_rect(data = log_ranges, 
            aes(xmin = max_log, xmax = Inf, ymin = -Inf, ymax = Inf), 
            fill = "grey", alpha = 0.2) +  # Right side greyed out
  # Add ribbons for confidence intervals
  geom_ribbon(data = preds, 
              aes(x = x, ymin = conf.low, ymax = conf.high, fill = group), 
              alpha = 0.2) +
  # Add predicted lines
  geom_line(data = preds, aes(x = x, y = predicted, color = group), size = 1) +  
  geom_abline(intercept = 0, 
              slope = 1, 
              color = "black", 
              size = 1) +
  facet_wrap(~ Taxa) +  
  labs(x = "log(size at t0)", 
       y = "log(size at t1)") +
  theme_bw() +
  theme(
    legend.title = element_blank(),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank()
  ) +
  
  scale_color_manual(values = c("2022" = "cadetblue", "2023" = "#D55E00")) +
  scale_fill_manual(values = c("2022" = "cadetblue", "2023" = "#D55E00"), guide = FALSE) +  # For ribbons only
  scale_x_continuous(breaks = seq(2, 10, by = 1), limits = c(2, 10)) +
  scale_y_continuous(breaks = seq(2, 10, by = 2), limits = c(2, 10)) +
  geom_segment(data = zero_crossings, 
               aes(x = x, xend = x, y = 0, yend = x, color = group), 
               linetype = "dashed", size = 1)



```

This plot shows model predictions for size at t depending on size at t previous. The dashed line represents the no-growth line where size t = size t previous. The vertical lines represent the size at which corals stopped to grow for each year (when the regression lines cross the no-growth line). We can see that for most taxa, the size at which corals stopped to grow are very similar between years but also between taxa (\~ 7 on log scale = 1100 cm2 = 37 cm in diameter on average).

We can also calculate the growth rates (slopes) of each taxon for each year:

```{r}
# Calculate growth rates
growth_rates <- preds |>
  group_by(Taxa, group) |>
  do(tidy(lm(predicted ~ x, data = .))) |>
  filter(term == "x") |>
  dplyr::select(Taxa, group, slope = estimate, std.error = std.error)

# View the growth rates
growth_rates
```

All growth rates are \< 1 with positive intercepts meaning that all taxa show decelerating growth which is consistent with allometric growth.


# 2. Effect of bleaching severity, size and environmental variables on survival and growth

## 2.1. Survival

```{r}
#| label: ReadData survival
#| output: false
#| echo: false
#| eval: true
#| warning: false
#| message: false
surv2 <- read_csv("C:/Users/jc815475/OneDrive - James Cook University/PhD data/3_Coral growth/Data/CSV/survival_transition2.csv", trim_ws=TRUE)
```

```{r}
#| label: survival prep
#| output: false
#| echo: false
#| eval: true
#| warning: false
#| message: false
surv_2 <- surv2 |> 
  mutate(Reef = factor(Reef),
         Site = factor(Site),
         Plot = factor(Plot),
         Colony = factor(Colony),
         Habitat = factor(Habitat),
         Depth = factor(Depth),
         Cross_shelf = factor(Cross_shelf),
         Taxa = factor(Taxa))|> 
  filter(Taxa %in% c("Abra", "Acor", "Pdam", "Pver", "Shys"))

surv_2 <- surv_2 |> rename(Survival = "Survival_2",
                           log_size = "log_2022")

surv_2_sub <- surv_2 |> 
  filter(!Bleaching_scores == "6")

surv_2_sub <- surv_2_sub |> 
  mutate(Bleaching_scores = factor(Bleaching_scores),
         bl_cat1 = fct_recode(Bleaching_scores,
                              `no bleaching` = "1",
                              `minor bleaching` = "2",
                              `major bleaching` = "3",
                              `fully bleached` = "4",
                              `recent partial mortality` = "5"),
         bl_cat2 = fct_recode(Bleaching_scores,
                              `no bleaching` = "1",
                              `partial bleaching` = "2",
                              `partial bleaching` = "3",
                              `fully bleached` = "4",
                              `recent partial mortality` = "5"))

```

Append environmental variables:
```{r}
#| label: model env data prep
#| echo: false
#| results: false
#| warning: false
#| message: false

envdata <- read_csv("C:/Users/jc815475/OneDrive - James Cook University/PhD data/3_Coral growth/Data/CSV/envdata_2022-2023.csv", trim_ws=TRUE)
env <- envdata |>
  select(Site_code, ubedmean, Secchi_mean, temp_mean)
env <- env |> rename(Site = Site_code)

env <- env |> 
  mutate(
    ubedmean_z = scale(ubedmean)[,1],
    Secchi_mean_z = scale(Secchi_mean)[,1],
    temp_mean_z = scale(temp_mean)[,1]
  )

## merge datasets
surv_env_dat <- merge(surv_2_sub, env, by = "Site", all.x = TRUE)

surv_env_dat <- surv_env_dat |> 
  mutate(Taxa = factor(Taxa)) |> 
  mutate(Site_code = str_sub(Site, 1, 8))


## filter for NAs
surv_env_dat <- surv_env_dat[complete.cases(surv_env_dat[, c("ubedmean_z", "Secchi_mean_z", "temp_mean")]), ]

surv_env_dat$bl_cat2 <- relevel(surv_env_dat$bl_cat2, ref = "no bleaching")
surv_env_dat$bl_cat1 <- relevel(surv_env_dat$bl_cat1, ref = "no bleaching")

surv_env_dat |> 
  group_by(Taxa) |> 
  summarise(total_count = n())#lowest sample size is Pdam with 208 colonies
```

### 2.1.1. Model fitting and selection

We formulate the hypothesis that water flow (by lowering water temperature) and clarity (by lowering light intensity) could help bleached corals survive better.

*Note: prior to this analysis, we tested whether initial size had a significant influence on survival by running model selection by comparing AIC of three models per taxon: 1) initial size interacting with bleaching severity, 2) initial size + bleaching severity and 3) just bleaching severity. For all morphotaxa, the best-fit model had initial size, interacting with bleaching severity or not.See Supplementary analyses section below*


Testing different random effect structures:

```{r}
#| label: randon effect selection survival Abra
Abra_s0 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Reef/Site_code/Plot),
              data = surv_env_dat[surv_env_dat$Taxa == "Abra",],
              family = "binomial")

Abra_s1 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Reef/Site_code),
              data = surv_env_dat[surv_env_dat$Taxa == "Abra",],
              family = "binomial")

Abra_s2 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Site_code/Plot),
              data = surv_env_dat[surv_env_dat$Taxa == "Abra",],
              family = "binomial")

Abra_s3 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Site_code),
              data = surv_env_dat[surv_env_dat$Taxa == "Abra",],
              family = "binomial")

```

```{r}
#| label: randon effect selection survival Abra AIC
AIC(Abra_s0, Abra_s1, Abra_s2, Abra_s3) #best structure is 1|Site_code for Abra
```

```{r}
#| label: randon effect selection survival Acor
Acor_s0 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Reef/Site_code/Plot),
              data = surv_env_dat[surv_env_dat$Taxa == "Acor",],
              family = "binomial")

Acor_s1 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Reef/Site_code),
              data = surv_env_dat[surv_env_dat$Taxa == "Acor",],
              family = "binomial")

Acor_s2 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Site_code/Plot),
              data = surv_env_dat[surv_env_dat$Taxa == "Acor",],
              family = "binomial")

Acor_s3 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Site_code),
              data = surv_env_dat[surv_env_dat$Taxa == "Acor",],
              family = "binomial")

```

```{r}
#| label: randon effect selection survival Acor AIC
AIC(Acor_s0, Acor_s1, Acor_s2, Acor_s3) #best structure is 1|Site_code for Acor
```

```{r}
#| label: randon effect selection survival Pdam
Pdam_s0 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Reef/Site_code/Plot),
              data = surv_env_dat[surv_env_dat$Taxa == "Pdam",],
              family = "binomial")

Pdam_s1 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Reef/Site_code),
              data = surv_env_dat[surv_env_dat$Taxa == "Pdam",],
              family = "binomial")

Pdam_s2 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Site_code/Plot),
              data = surv_env_dat[surv_env_dat$Taxa == "Pdam",],
              family = "binomial")

Pdam_s3 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Site_code),
              data = surv_env_dat[surv_env_dat$Taxa == "Pdam",],
              family = "binomial")

```

```{r}
#| label: randon effect selection survival Pdam AIC
AIC(Pdam_s0, Pdam_s1, Pdam_s2, Pdam_s3) #best structure is 1|Site_code for Pdam
```

```{r}
#| label: randon effect selection survival Pver
Pver_s0 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Reef/Site_code/Plot),
              data = surv_env_dat[surv_env_dat$Taxa == "Pver",],
              family = "binomial")

Pver_s1 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Reef/Site_code),
              data = surv_env_dat[surv_env_dat$Taxa == "Pver",],
              family = "binomial")

Pver_s2 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Site_code/Plot),
              data = surv_env_dat[surv_env_dat$Taxa == "Pver",],
              family = "binomial")

Pver_s3 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Site_code),
              data = surv_env_dat[surv_env_dat$Taxa == "Pver",],
              family = "binomial")

```

```{r}
#| label: randon effect selection survival Pver AIC
AIC(Pver_s0, Pver_s1, Pver_s2, Pver_s3) #best structure is 1|Site_code for Pver
```

```{r}
#| label: randon effect selection survival Shys
Shys_s0 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Reef/Site_code/Plot),
              data = surv_env_dat[surv_env_dat$Taxa == "Shys",],
              family = "binomial")

Shys_s1 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Reef/Site_code),
              data = surv_env_dat[surv_env_dat$Taxa == "Shys",],
              family = "binomial")

Shys_s2 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Site_code/Plot),
              data = surv_env_dat[surv_env_dat$Taxa == "Shys",],
              family = "binomial")

Shys_s3 <- glmmTMB(Survival ~ log_size : bl_cat1  +
                (1|Site_code),
              data = surv_env_dat[surv_env_dat$Taxa == "Shys",],
              family = "binomial")

```

```{r}
#| label: randon effect selection survival Shys AIC
AIC(Shys_s0, Shys_s1, Shys_s2, Shys_s3) #best structure is 1|Site_code/Plot for Shys so will need to run model selection separately
```

Based on the best base models and best random effect structure for each taxon, We create a function that adapts dynamically to each taxon's needs. Rather than writing five separate functions to apply model selection to test all environmental variables, we can parametrise one flexible function using a simple configuration list per taxon.

First let's create a configuration list per taxon:

```{r}
#|label: create configuration lists

taxon_configs <- list(
  Abra = list(
    base_formula = Survival ~ log_size : bl_cat1,
    random_effects = "(1 | Site_code)"
  ),
  Acor = list(
    base_formula = Survival ~ log_size + bl_cat1,
    random_effects = "(1 | Site_code)"
  ),
  Pdam = list(
    base_formula = Survival ~ log_size + bl_cat1,
    random_effects = "(1 | Site_code)"
  ),
  Pver = list(
    base_formula = Survival ~ log_size + bl_cat1,
    random_effects = "(1 | Site_code)"
  ),
  Shys = list(
    base_formula = Survival ~ log_size : bl_cat1,
    random_effects = "(1 | Site_code/Plot)"
  )
)

```

Then write a flexible model-fitting function:

```{r}
#|label: create flexible function to fit models

fit_survival_models_strategic <- function(taxon, data, config) {
  df <- data |> dplyr::filter(Taxa == taxon)
  
  base_formula <- config$base_formula
  re_structure <- config$random_effects
  
  # Extract base fixed effects as a string
  base_fixed_str <- paste(deparse(base_formula[[3]]), collapse = "")
  
  # Combine base and added terms as a formula string
  make_full_formula <- function(extra_terms) {
    # Combine fixed effects and random effects
    formula_str <- paste0("Survival ~ ", base_fixed_str, 
                          if (!is.null(extra_terms)) paste0(" + ", extra_terms) else "", 
                          " + ", re_structure)
    as.formula(formula_str)
  }

  # Define all model formulas
  surv_formulas <- list(
    full           = make_full_formula("Secchi_mean_z * bl_cat1 + ubedmean_z * bl_cat1 + temp_mean_z * bl_cat1"),
    null           = make_full_formula(NULL),
    
    no_clarity     = make_full_formula("ubedmean_z * bl_cat1 + temp_mean_z * bl_cat1"),
    no_velocity    = make_full_formula("Secchi_mean_z * bl_cat1 + temp_mean_z * bl_cat1"),
    no_temp        = make_full_formula("Secchi_mean_z * bl_cat1 + ubedmean_z * bl_cat1"),
    
    only_clarity   = make_full_formula("Secchi_mean_z * bl_cat1"),
    only_velocity  = make_full_formula("ubedmean_z * bl_cat1"),
    only_temp      = make_full_formula("temp_mean_z * bl_cat1")
  )

  # Fit the models
  surv_models <- lapply(surv_formulas, function(f) {
    glmmTMB(f, data = df, family = binomial)
  })

  return(surv_models)
}




```

Run the models for each taxon using their config:

```{r}
#|label: run models depending on configs

taxa_list <- names(taxon_configs)

surv_model_results <- lapply(taxa_list, function(taxon) {
  fit_survival_models_strategic(taxon, surv_env_dat, taxon_configs[[taxon]])
})
names(surv_model_results) <- taxa_list

```

then we extract AIC to select best models per taxon:

```{r}
#|label: extract AIC

surv_aic_results <- lapply(surv_model_results, function(taxon_models) {
  sapply(taxon_models, function(m) {
    if (!is.null(m) && !inherits(m, "try-error")) AIC(m) else NA
  })
})

for (taxon in names(surv_aic_results)) {
  cat("\n--- AIC Results for", taxon, "---\n")
  print(sort(surv_aic_results[[taxon]]))  # sorts from best (lowest AIC) to worst
}
```

### 2.1.2. Model summary and diagnostics

Acropora staghorn

```{r}
#|label: summary survival best-fit model Abra
summary(surv_model_results$Abra$null)
```

Model diagnostics:

```{r}
#|label: simulated residuals Abra
Abra_sim_res <- simulateResiduals(surv_model_results$Abra$null, plot = TRUE)
```

```{r}
#|label: overdispersion Abra
testDispersion(Abra_sim_res)
```

```{r}
#|label: multicollinearity Abra
# library(car)
# car::vif(model_results$Abra$s0) no need
```

Acropora corymbose

```{r}
#|label: summary survival best-fit model Acor
summary(surv_model_results$Acor$null)
```

Best-fit model for *A. corymbose* includes additive fixed effects of initial size and bleaching severity, the survival probability of colonies in the bleaching category "recent partial mortality" is significantly different from "no bleaching"

Model diagnostics:

```{r}
#|label: simulated residuals Acor
Acor_sim_res <- simulateResiduals(surv_model_results$Acor$null, plot = TRUE)
```

```{r}
#|label: overdispersion Acor
testDispersion(Acor_sim_res)
```

```{r}
#|label: multicollinearity Acor

#car::vif(model_results$Acor$s0g)
```

Pocillopora damicornis

```{r}
#|label: summary survival best-fit model Pdam
summary(surv_model_results$Pdam$null)
```

Model diagnostics:

```{r}
#|label: simulated residuals Pdam
Pdam_sim_res <- simulateResiduals(surv_model_results$Pdam$null, plot = TRUE)
```

```{r}
#|label: overdispersion Pdam
testDispersion(Pdam_sim_res)
```

For *P. damicornis*, the best-fit model includes initial size with a positive effect on survival and bleaching severity. Only the effect of 'recent partial mortality' category is significantly different from the 'no bleaching' category.

Pocillopora verrucosa

```{r}
#|label: summary survival best-fit model Pver
summary(surv_model_results$Pver$null)
```

Model diagnostics:

```{r}
#|label: simulated residuals Pver
Pver_sim_res <- simulateResiduals(surv_model_results$Pver$null, plot = TRUE)
```

```{r}
#|label: overdispersion Pver
testDispersion(Pver_sim_res)
```

For *P. verrucosa*, the best-fit model includes initial size with a positive effect on survival and bleaching severity with significant negative effect. The effect of bleaching severity 'major bleaching', 'fully bleached' and 'recent partial mortality' are significantly different from 'no bleaching'.

Seriatopora hystrix

```{r}
#|label: summary survival best-fit model Shys
summary(surv_model_results$Shys$only_temp)
```

Model diagnostics:

```{r}
#|label: simulated residuals Shys
Shys_sim_res <- simulateResiduals(surv_model_results$Shys$null, plot = TRUE)
```

```{r}
#|label: overdispersion Shys
testDispersion(Shys_sim_res)
```

For *S. hystrix*, the best-fit model includes initial size interacting with bleaching severity and temperature interacting with bleaching severity but this interaction isn't significant.

### 2.1.3. Summary figures

#### 2.1.3.a. Figure 4: Acropora staghorn

```{r}
#| label: Abra surv preds
#| echo: false
Abra_best_model <- model_results$Abra$null
Abra_s_preds <- ggpredict(Abra_best_model, terms = c("log_size [all]", "bl_cat1"))
```

```{r}
#| label: Abra surv figure
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8

colours_sev = c("#F0E442",  "#E69F00", "#D55E00", "red", "black")


ggplot() +
  geom_line(data = Abra_s_preds, 
            aes(x = x, y = predicted, color = group), 
            size = 1) +
  geom_ribbon(data = Abra_s_preds,
              aes(x = x,
                  ymin = conf.low,
                  ymax = conf.high,
                  fill = group),
              alpha = 0.2,
              inherit.aes = FALSE,
              show.legend = FALSE) +
  labs(x = "Initial size (log)",
       y = "Survival probability",
       color = "") +
  scale_color_manual(values = colours_sev) +
  scale_fill_manual(values = colours_sev) +
  theme_bw() +
  theme(
    legend.position = "right")


```

For *A. staghorn*, the best-fit model is the null model which includes only the interaction between initial size and bleaching severity.

#### 2.1.3.b. Figure 5: Acropora corymbose

```{r}
#| label: Acor surv preds
#| echo: false
#| message: false
#| warning: false
Acor_best_model <- model_results$Acor$null

Acor_s_preds <- ggpredict(Acor_best_model, terms = "log_size [all], ", type = "fixed")

Acor_s_preds2 <- ggpredict(Acor_best_model, terms = "bl_cat1 [no bleaching], ", type = "fixed")
Acor_s_preds3 <- ggpredict(Acor_best_model, terms = "bl_cat1 [minor bleaching], ", type = "fixed")
Acor_s_preds4 <- ggpredict(Acor_best_model, terms = "bl_cat1 [major bleaching], ", type = "fixed")
Acor_s_preds5 <- ggpredict(Acor_best_model, terms = "bl_cat1 [fully bleached], ", type = "fixed")
Acor_s_preds6 <- ggpredict(Acor_best_model, terms = "bl_cat1 [recent partial mortality], ", type = "fixed")
Acor_s_preds1 <- rbind(Acor_s_preds2, Acor_s_preds3, Acor_s_preds4, Acor_s_preds5, Acor_s_preds6)


```

```{r}
#| label: Acor surv figure
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8

ggplot() +
  geom_line(data = Acor_s_preds, 
            aes(x = x, y = predicted), 
            size = 1) +
  # geom_point(data = surv_env_dat[surv_env_dat$Taxa == "Acor",],
  #            aes(x = log_size,
  #                y = Survival),
  #            size = 2,
  #            alpha = 0.2,
  #            pch=21,
  #            position = position_jitter(width = 0.1, height = 0.1)) +
  geom_ribbon(data = Acor_s_preds,
              aes(x = x,
                  ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.2,
              inherit.aes = FALSE,
              show.legend = FALSE) +
  labs(x = "Initial size (log)",
       y = "Survival probability") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.25),
                     limits = c(0, 1)) +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12))


ggplot(data = Acor_s_preds1, aes(x = x, y = predicted, fill = x)) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.1, linewidth = 0.8) +
  geom_point(size = 5, pch = 21, color = "black") +  
  scale_fill_manual(values = colours_sev, name = "Bleaching Severity") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.25), limits = c(0, 1)) +
  theme_bw() +
  labs(x = "", y = "Survival Probability") +
  theme(
    legend.position = "none",
    axis.text.x = element_blank()
  ) +
  scale_x_discrete(limits = levels(Acor_s_preds1$x))
```

For *A. corymbose*, the best-fit model only includes bleaching severity with an increasing negative effect on survival and initial size with a significant positive effect on survival probability. The effect of bleaching severity categories 'fully bleached' and 'recent partial mortality' are significantly different from the 'no bleaching' category.

#### 2.1.3.c. Figure 6: Pocillopora damicornis

```{r}
#| label: Pdam surv preds
#| echo: false
#| message: false
#| warning: false

Pdam_best_model <- model_results$Pdam$null

Pdam_s_preds1 <- ggpredict(Pdam_best_model, terms = "log_size [all], ", type = "fixed")

Pdam_s_preds2 <- ggpredict(Pdam_best_model, terms = "bl_cat1 [no bleaching], ", type = "fixed")
Pdam_s_preds3 <- ggpredict(Pdam_best_model, terms = "bl_cat1 [minor bleaching], ", type = "fixed")
Pdam_s_preds4 <- ggpredict(Pdam_best_model, terms = "bl_cat1 [fully bleached], ", type = "fixed")
Pdam_s_preds5 <- ggpredict(Pdam_best_model, terms = "bl_cat1 [recent partial mortality], ", type = "fixed")
Pdam_s_preds <- rbind(Pdam_s_preds2, Pdam_s_preds3, Pdam_s_preds4, Pdam_s_preds5)
```

```{r}
#| label: Pdam surv figure
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8

colours_sev1 = c("#F0E442",  "#E69F00", "red", "black")

ggplot() +
  geom_line(data = Pdam_s_preds1, 
            aes(x = x, y = predicted), 
            size = 1) +
  geom_ribbon(data = Pdam_s_preds1,
              aes(x = x,
                  ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.2,
              inherit.aes = FALSE,
              show.legend = FALSE) +
  labs(x = "Initial size (log)",
       y = "Survival probability") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.25),
                     limits = c(0, 1)) +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12))

ggplot(data = Pdam_s_preds, aes(x = x, y = predicted, fill = x)) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.1, linewidth = 0.8) +
  geom_point(size = 5, pch = 21, color = "black") +  
  scale_fill_manual(values = colours_sev1, name = "Bleaching Severity") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.25), limits = c(0, 1)) +
  theme_bw() +
  labs(x = "", y = "Survival Probability") +
  theme(
    legend.position = "none",
    axis.text.x = element_blank()
  ) +
  scale_x_discrete(limits = levels(Pdam_s_preds$x))

```

For *P. damicornis*, the best-fit model includes initial size with a positive effect on survival and bleaching severity. Only the effect of 'recent partial mortality' category is significantly different from the 'no bleaching' category.

#### 2.1.3.d. Figure 7: Pocillopora verrucosa

```{r}
#| label: Pver surv preds
#| echo: false
#| message: false
#| warning: false

Pver_best_model <- model_results$Pver$null
Pver_s_preds1 <- ggpredict(Pver_best_model, terms = "log_size [all]", type = "fixed")
Pver_s_preds2 <- ggpredict(Pver_best_model, terms = "bl_cat1 [no bleaching], ", type = "fixed")
Pver_s_preds3 <- ggpredict(Pver_best_model, terms = "bl_cat1 [minor bleaching], ", type = "fixed")
Pver_s_preds4 <- ggpredict(Pver_best_model, terms = "bl_cat1 [major bleaching], ", type = "fixed")
Pver_s_preds5 <- ggpredict(Pver_best_model, terms = "bl_cat1 [fully bleached], ", type = "fixed")
Pver_s_preds6 <- ggpredict(Pver_best_model, terms = "bl_cat1 [recent partial mortality], ", type = "fixed")
Pver_s_preds <- rbind(Pver_s_preds2, Pver_s_preds3, Pver_s_preds4, Pver_s_preds5, Pver_s_preds6)
```

```{r}
#| label: Pver surv figure
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8

colours_sev = c("#F0E442",  "#E69F00", "#D55E00", "red", "black")

ggplot() +
  geom_line(data = Pver_s_preds1, 
            aes(x = x, y = predicted), 
            size = 1) +
  geom_ribbon(data = Pver_s_preds1,
              aes(x = x,
                  ymin = conf.low,
                  ymax = conf.high),
              alpha = 0.2,
              inherit.aes = FALSE,
              show.legend = FALSE) +
  labs(x = "Initial size (log)",
       y = "Survival probability") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.25),
                     limits = c(0, 1)) +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12))

ggplot(data = Pver_s_preds, aes(x = x, y = predicted, fill = x)) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.1, linewidth = 0.8) +
  geom_point(size = 5, pch = 21, color = "black") +  
  scale_fill_manual(values = colours_sev, name = "Bleaching Severity") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.25), limits = c(0, 1)) +
  theme_bw() +
  labs(x = "", y = "Survival Probability") +
  theme(
    legend.position = "none",
    axis.text.x = element_blank()
  ) +
  scale_x_discrete(limits = levels(Pver_s_preds$x))
```

For *P. verrucosa*, the best-fit model includes initial size with a positive effect on survival and bleaching severity with increasing negative effect. The effect of bleaching severity 'fully bleached' and 'recent partial mortality' are significantly different from 'no bleaching'.

#### 2.1.3.e. Figure 8: Seriatopora hystrix

```{r}
#| label: Shys surv preds
#| echo: false
shys_best_model <- model_results$Shys$only_temp
Shys_s_preds <- ggpredict(shys_best_model, terms = c("log_size [all]", "bl_cat1"), type = "fixed") #average over temperature

Shys_s_preds1 <- ggpredict(shys_best_model, terms = c("temp_mean_z [all]"), type = "fixed") #average over bleaching categories
```

```{r}
#| label: Shys surv figure1
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8

colours_sev = c("#F0E442",  "#E69F00", "#D55E00", "red", "black")

ggplot() +
  geom_line(data = Shys_s_preds, 
            aes(x = x, y = predicted, color = group), 
            size = 1) +
  geom_ribbon(data = Shys_s_preds,
              aes(x = x,
                  ymin = conf.low,
                  ymax = conf.high,
                  fill = group),
              alpha = 0.2,
              inherit.aes = FALSE,
              show.legend = FALSE) +
  labs(x = "Initial size (log)",
       y = "Survival probability",
       color = "") +
  scale_color_manual(values = colours_sev) +
  scale_fill_manual(values = colours_sev) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.25),
                     limits = c(0, 1)) +
  theme_bw() +
  theme(
    legend.position = "right")

ggplot(Shys_s_preds1, aes(x = x, y = predicted)) +
  geom_line(size = 1.2, color = "#714674") +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), fill = "#FF99FF", alpha = 0.3) +
  labs(
    x = "Standardised Temperature",
    y = "Survival Probability"
  ) +
  theme_bw()
```

```{r}
#| label: Shys surv figure2
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8

# ggplot() +
#   geom_line(data = Shys_s0a_preds1, 
#             aes(x = x, y = predicted), 
#             size = 1) +
#   geom_ribbon(data = Shys_s0a_preds1,
#               aes(x = x,
#                   ymin = conf.low,
#                   ymax = conf.high),
#               alpha = 0.2,
#               inherit.aes = FALSE,
#               show.legend = FALSE) +
#   labs(x = "ubedmean",
#        y = "Survival probability",
#        color = "") +
#   scale_y_continuous(breaks = seq(0, 1, by = 0.25),
#                      limits = c(0, 1)) +
#   theme_bw() +
#   theme(
#     legend.position = "right")
```

```{r}
#| label: Shys surv figure3
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8
# 
# ggplot() +
#   geom_line(data = Shys_s0c_preds2, 
#             aes(x = x, y = predicted), 
#             size = 1) +
#   geom_ribbon(data = Shys_s0c_preds2,
#               aes(x = x,
#                   ymin = conf.low,
#                   ymax = conf.high),
#               alpha = 0.2,
#               inherit.aes = FALSE,
#               show.legend = FALSE) +
#   labs(x = "Secchi mean",
#        y = "Survival probability",
#        color = "") +
#   scale_y_continuous(breaks = seq(0, 1, by = 0.25),
#                      limits = c(0, 1)) +
#   theme_bw() +
#   theme(
#     legend.position = "right")
```

In summary, no environmental variables were retained in the best-fit models for Survival.

## 2.2 Growth

```{r}
#| label: ReadData growth
#| output: false
#| echo: false
#| eval: true
#| warning: false
#| message: false
grow2 <- read_csv("C:/Users/jc815475/OneDrive - James Cook University/PhD data/3_Coral growth/Data/CSV/growth_transition2.csv", trim_ws=TRUE)
```

```{r}
#| label: growth prep
#| output: false
#| echo: false
#| eval: true
#| warning: false
#| message: false
grow_2 <- grow2 |> 
  mutate(Reef = factor(Reef),
         Site = factor(Site),
         Site_code = str_sub(Site, 1, 8),
         Plot = factor(Plot),
         Colony = factor(Colony),
         Habitat = factor(Habitat),
         Depth = factor(Depth),
         Cross_shelf = factor(Cross_shelf),
         Taxa = factor(Taxa))|> 
  filter(Taxa %in% c("Abra", "Acor", "Pdam", "Pver", "Shys"))

grow_2 <- grow_2 |> 
  mutate(bl_cat1 = case_when(
    is.na(Bleaching_scores)  ~ "no bleaching",
    Bleaching_scores == 1   ~ "no bleaching",
    Bleaching_scores == 2   ~ "minor bleaching",
    Bleaching_scores == 3   ~ "major bleaching",
    Bleaching_scores == 4   ~ "fully bleached",
    Bleaching_scores == 5   ~ "recent partial mortality"
  ))

grow_2 <- grow_2 |> 
  mutate(bl_cat2 = fct_recode(bl_cat1,
                              `no bleaching` = "no bleaching",
                              `partial bleaching` = "minor bleaching",
                              `partial bleaching` = "major bleaching",
                              `fully bleached` = "fully bleached",
                              `recent partial mortality` = "recent partial mortality"))

grow_2 <- grow_2 |> 
  mutate(bl_cat2 = factor(bl_cat2, levels = c("no bleaching", "partial bleaching", "fully bleached", "recent partial mortality")))

grow_2 <- grow_2 |> 
  mutate(bl_cat1 = factor(bl_cat1, levels = c("no bleaching", "minor bleaching", "major bleaching", "fully bleached", "recent partial mortality")))
```

```{r}
#| label: data prep growth2
#| echo: false
#| warning: false
#| message: false

# envdata <- read_csv("C:/Users/jc815475/OneDrive - James Cook University/PhD data/3_Coral growth/Data/CSV/All_environmental_data_EcoRRAP_sites.csv", trim_ws=TRUE)
# 
# envdata <- envdata |> 
#   mutate(Zone = factor(Zone),
#          Site.code = factor(Site.code))
# envdata <- envdata |> 
#   filter(! Zone == "F") |> 
#   filter(! Zone == "B")
# envdata <- envdata |> 
#   mutate(Reef = str_sub(Site.code, 1, 4)) |> 
#   mutate(Reef = factor(Reef))
# envdat <- envdata |> 
#   filter(Reef %in% c("OCCH", "OCDA", "OCLB", "PAOR", "PAPE"))
# 
# env <- envdat |> 
#   select(Site.code, ubedmean, Secchi_mean, temp_mean)
# env <- env |> rename(Site = Site.code)

envdata <- read_csv("C:/Users/jc815475/OneDrive - James Cook University/PhD data/3_Coral growth/Data/CSV/envdata_2022-2023.csv", trim_ws=TRUE)
env <- envdata |>
  select(Site_code, ubedmean, Secchi_mean, temp_mean)
env <- env |> rename(Site = Site_code)

env <- env |> 
  mutate(
    ubedmean_z = scale(ubedmean)[,1],
    Secchi_mean_z = scale(Secchi_mean)[,1],
    temp_mean_z = scale(temp_mean)[,1]
  )

## merge datasets
grow_env_dat <- merge(grow_2, env, by = "Site", all.x = TRUE)

grow_env_dat <- grow_env_dat |> 
  mutate(Taxa = factor(Taxa)) |> 
  mutate(Site_code = str_sub(Site, 1, 8))

## filter for NAs
grow_env_dat <- grow_env_dat[complete.cases(grow_env_dat[, c("ubedmean_z", "Secchi_mean_z", "temp_mean_z")]), ]

grow_env_dat |> 
  group_by(Taxa) |> 
  summarise(total_count = n())#lowest sample size is Pdam with 89 colonies
```

We fitted a separate model for each taxon, comparing log-transformed size at t0 (2022 pre-bleaching) versus size at t1 (2023). We selected three environmental variables based on exploratory data analyses and previous works on factors affecting coral growth rates in the context of bleaching. The environmental variables that were included alongside log-transformed size at t0 were water velocity, turbidity and temperature.

### 2.2.1. Model fitting and selection

Test random effect structure

```{r}
#| label: randon effect selection growth Abra
Abra_g0 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Reef/Site_code/Plot),
              data = grow_env_dat[grow_env_dat$Taxa == "Abra",])

Abra_g1 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Reef/Site_code),
              data = grow_env_dat[grow_env_dat$Taxa == "Abra",])

Abra_g2 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Site_code/Plot),
              data = grow_env_dat[grow_env_dat$Taxa == "Abra",])

Abra_g3 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Site_code),
              data = grow_env_dat[grow_env_dat$Taxa == "Abra",])

```

```{r}
#| label: randon effect selection growth Abra AIC
AIC(Abra_g0, Abra_g1, Abra_g2, Abra_g3) #best structure is 1|Site_code for Abra
```

```{r}
#| label: randon effect selection growth Acor
Acor_g0 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Reef/Site_code/Plot),
              data = grow_env_dat[grow_env_dat$Taxa == "Acor",])

Acor_g1 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Reef/Site_code),
              data = grow_env_dat[grow_env_dat$Taxa == "Acor",])

Acor_g2 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Site_code/Plot),
              data = grow_env_dat[grow_env_dat$Taxa == "Acor",])

Acor_g3 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Site_code),
              data = grow_env_dat[grow_env_dat$Taxa == "Acor",])

```

```{r}
#| label: randon effect selection growth Acor AIC
AIC(Acor_g0, Acor_g1, Acor_g2, Acor_g3) #best structure is 1|Site_code for Acor
```

```{r}
#| label: randon effect selection growth Pdam
Pdam_g0 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Reef/Site_code/Plot),
              data = grow_env_dat[grow_env_dat$Taxa == "Pdam",])

Pdam_g1 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Reef/Site_code),
              data = grow_env_dat[grow_env_dat$Taxa == "Pdam",])

Pdam_g2 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Site_code/Plot),
              data = grow_env_dat[grow_env_dat$Taxa == "Pdam",])

Pdam_g3 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Site_code),
              data = grow_env_dat[grow_env_dat$Taxa == "Pdam",])

```

```{r}
#| label: randon effect selection growth Pdam AIC
AICc(Pdam_g0, Pdam_g1, Pdam_g2, Pdam_g3) #best structure is 1|Site_code for Pdam
```

```{r}
#| label: randon effect selection growth Pver
Pver_g0 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Reef/Site_code/Plot),
              data = grow_env_dat[grow_env_dat$Taxa == "Pver",])

Pver_g1 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Reef/Site_code),
              data = grow_env_dat[grow_env_dat$Taxa == "Pver",])

Pver_g2 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Site_code/Plot),
              data = grow_env_dat[grow_env_dat$Taxa == "Pver",])

Pver_g3 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Site_code),
              data = grow_env_dat[grow_env_dat$Taxa == "Pver",])

```

```{r}
#| label: randon effect selection growth Pver AIC
AICc(Pver_g0, Pver_g1, Pver_g2, Pver_g3) #best structure is 1|Site_code for Pver
```

```{r}
#| label: randon effect selection growth Shys
Shys_g0 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Reef/Site_code/Plot),
              data = grow_env_dat[grow_env_dat$Taxa == "Shys",])

Shys_g1 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Reef/Site_code),
              data = grow_env_dat[grow_env_dat$Taxa == "Shys",])

Shys_g2 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Site_code/Plot),
              data = grow_env_dat[grow_env_dat$Taxa == "Shys",])

Shys_g3 <- glmmTMB(log_2023 ~ log_2022 + log_2022:bl_cat1  +
                (1|Site_code),
              data = grow_env_dat[grow_env_dat$Taxa == "Shys",])

```

```{r}
#| label: randon effect selection growth Shys AIC
AICc(Shys_g0, Shys_g1, Shys_g2, Shys_g3) #best structure is 1|Site_code for Shys
```

Random effect structure is (1\|Site_code) for all taxa. We are going to test the different models below:

Full model: log_2023 \~ log_2022 + log_2022:bl_cat2 + log_2022:Secchi_mean + log_2022:ubedmean + log_2022:temp_mean + log_2022:Secchi_mean:bl_cat + log_2022:ubedmean:bl_cat2 + log_2022:temp_mean:bl_cat2 + (1\|Site_code)

Intermediate models:

-   no water clarity: log_2023 \~ log_2022 + log_2022:bl_cat2 + log_2022:ubedmean + log_2022:temp_mean + log_2022:ubedmean:bl_cat2 + log_2022:temp_mean:bl_cat2 + (1\|Site_code)

-   no water velocity: log_2023 \~ log_2022 + log_2022:bl_cat2 + log_2022:Secchi_mean + log_2022:temp_mean + log_2022:Secchi_mean:bl_cat + log_2022:temp_mean:bl_cat2 + (1\|Site_code)

-   no temperature: log_2023 \~ log_2022 + log_2022:bl_cat2 + log_2022:Secchi_mean + log_2022:ubedmean + log_2022:Secchi_mean:bl_cat + log_2022:ubedmean:bl_cat2 + (1\|Site_code)

-   only water clarity: log_2023 \~ log_2022 + log_2022:bl_cat2 + log_2022:Secchi_mean + log_2022:Secchi_mean:bl_cat2 + (1\|Site_code)

-   only water velocity: log_2023 \~ log_2022 + log_2022:bl_cat2 + log_2022:ubedmean + log_2022:ubedmean:bl_cat2 + (1\|Site_code)

-   only temperature: log_2023 \~ log_2022 + log_2022:bl_cat2 + log_2022:temp_mean + log_2022:temp_mean:bl_cat2 + (1\|Site_code)

-   no interaction bleaching x environment: log_2023 \~ log_2022 + log_2022:bl_cat2 + log_2022:Secchi_mean + log_2022:ubedmean + log_2022:temp_mean + (1\|Site_code)

Null model: log_2023 \~ log_2022 + log_2022:bl_cat2 + (1 \| Site_code)

```{r}
#| label: model 3.1 formulas
#| echo: false
#| message: false
#| warning: false 

growth_formulas <- list(
  full      = log_2023 ~ log_2022 + log_2022:bl_cat1 + log_2022:Secchi_mean_z + log_2022:ubedmean_z + log_2022:temp_mean_z +
              log_2022:Secchi_mean_z:bl_cat1 + log_2022:ubedmean_z:bl_cat1 + log_2022:temp_mean_z:bl_cat1 + (1 | Site_code),
  
  no_clarity = log_2023 ~ log_2022 + log_2022:bl_cat1 + log_2022:ubedmean_z + log_2022:temp_mean_z +
               log_2022:ubedmean_z:bl_cat1 + log_2022:temp_mean_z:bl_cat1 + (1 | Site_code),
  
  no_velocity = log_2023 ~ log_2022 + log_2022:bl_cat1 + log_2022:Secchi_mean_z + log_2022:temp_mean_z +
                log_2022:Secchi_mean_z:bl_cat1 + log_2022:temp_mean_z:bl_cat1 + (1 | Site_code),
  
  no_temp = log_2023 ~ log_2022 + log_2022:bl_cat1 + log_2022:Secchi_mean_z + log_2022:ubedmean_z +
            log_2022:Secchi_mean_z:bl_cat1 + log_2022:ubedmean_z:bl_cat1 + (1 | Site_code),
  
  only_clarity = log_2023 ~ log_2022 + log_2022:bl_cat1 + log_2022:Secchi_mean_z +
                 log_2022:Secchi_mean_z:bl_cat1 + (1 | Site_code),
  
  only_velocity = log_2023 ~ log_2022 + log_2022:bl_cat1 + log_2022:ubedmean_z +
                  log_2022:ubedmean_z:bl_cat1 + (1 | Site_code),
  
  only_temp = log_2023 ~ log_2022 + log_2022:bl_cat1 + log_2022:temp_mean_z +
              log_2022:temp_mean_z:bl_cat1 + (1 | Site_code),
  
  additive_env = log_2023 ~ log_2022 + log_2022:bl_cat1 + log_2022:Secchi_mean_z + log_2022:ubedmean_z + log_2022:temp_mean_z + (1 | Site_code),
  
  null = log_2023 ~ log_2022 + log_2022:bl_cat1 + (1 | Site_code)
)

```

```{r}
#| label: model 3.1 prep cont.
#| echo: false
#| message: false
#| warning: false 

#remove obvious outliers
grow_env_dat <- grow_env_dat |> filter(!Colony == "OCDA_LA2S_P4_2927")

taxa_list <- c("Abra", "Acor", "Pdam", "Pver", "Shys")
datasets <- lapply(taxa_list, function(taxon) {
  df <- grow_env_dat[grow_env_dat$Taxa == taxon, ]
  df$bl_cat1 <- relevel(df$bl_cat1, ref = "no bleaching")
  df$bl_cat2 <- relevel(df$bl_cat2, ref = "no bleaching")
  return(df)
})
names(datasets) <- taxa_list

```

```{r}
#| label: model 3.1 loop
#| echo: false
#| message: false
#| warning: false 

results <- list()

for (taxon_name in names(datasets)) {
  taxon_data <- datasets[[taxon_name]]
  taxon_results <- list()
  
  for (model_name in names(growth_formulas)) {
    formula <- growth_formulas[[model_name]]
    model <- glmmTMB(formula, data = taxon_data)
    taxon_results[[model_name]] <- model
  }
  
  results[[taxon_name]] <- taxon_results
}


```

```{r}
#| label: models 3.1 AIC
#| echo: false
aic_results <- lapply(results, function(taxon_models) {
  sapply(taxon_models, function(m) {
    if (!is.null(m) && !inherits(m, "try-error")) AICc(m) else NA
  })
})
```

```{r}
#| label: models 3.1 Print sorted AIC Results
#| echo: false

for (taxon in names(aic_results)) {
  cat("\n--- AICc Results for", taxon, "---\n")
  print(sort(aic_results[[taxon]]))  # sorts from best (lowest AIC) to worst
}
```

The model "only_temp" was the best-fit model for Abra, Pver and Shys. The "additive_env" model was the best-fit model for Acor and the null model was the best-fit model for Pdam.

```{r}
#| label: models 3.1 select best models
Abra_best_mod <- results$Abra$only_temp
Acor_best_mod <- results$Acor$additive_env
Pdam_best_mod <- results$Pdam$null
Pver_best_mod <- results$Pver$only_temp
Shys_best_mod <- results$Shys$only_temp
```

### 2.2.2. Model summary and diagnostics

```{r}
#|label: residuals growth Abra

best_models <- list(
  Abra = results$Abra$only_temp,
  Acor = results$Acor$additive_env,
  Pdam = results$Pdam$null,
  Pver = results$Pver$only_temp,
  Shys = results$Shys$only_temp
)


sim_residuals <- list()
dispersion_tests <- list()
collinearity_checks <- list()

for(taxon in names(best_models)) {
  cat("Running diagnostics for", taxon, "...\n")
  
  model <- best_models[[taxon]]
  
  # Simulate residuals and plot
  sim_res <- simulateResiduals(model, plot = TRUE)
  sim_residuals[[taxon]] <- sim_res
  
  # Test dispersion
  disp_test <- testDispersion(sim_res)
  dispersion_tests[[taxon]] <- disp_test
  
  # Check collinearity
  collin <- check_collinearity(model)
  collinearity_checks[[taxon]] <- collin
  
  print(disp_test)
  print(collin)
  cat("\n-----------------------\n")
}

```

High variance inflation factors (VIFs \> 90) were observed for higher-order interaction terms in Pver and Shys, likely due to low sample size in certain bleaching categories. We retained these terms to maintain model comparability, but interpret results from these interactions with caution.

### 2.2.3. Summary figures

#### 2.2.3.a. Figure 4: Acropora staghorn

```{r}
#| label: abra summary
#| echo: false
summary(Abra_best_mod)
```

The best-fit model for *A. staghorn* included water temperature interacting with initial size and bleaching severity. Only the categories 'fully bleached' and 'recent partial mortality' showed significance indicating that corals in these categories are significantly different from the 'no bleaching' category and have a stronger relationship with water temperature when considering initial size.

```{r}
#| label: Abra preds
#| echo: false
#| warning: false
#| message: false

# effect of bleaching severity
Abra_g_preds0 <- ggpredict(Abra_best_mod, terms = c("log_2022 [all]", "bl_cat1 [no bleaching]"), type = "fixed")
Abra_g_preds1 <- ggpredict(Abra_best_mod, terms = c("log_2022 [all]", "bl_cat1 [minor bleaching]"), type = "fixed")
Abra_g_preds2 <- ggpredict(Abra_best_mod, terms = c("log_2022 [all]", "bl_cat1 [major bleaching]"), type = "fixed")
Abra_g_preds3 <- ggpredict(Abra_best_mod, terms = c("log_2022 [all]", "bl_cat1 [fully bleached]"), type = "fixed")
Abra_g_preds4 <- ggpredict(Abra_best_mod, terms = c("log_2022 [all]", "bl_cat1 [recent partial mortality]"), type = "fixed")

Abra_g_predsa <- rbind(Abra_g_preds0, Abra_g_preds1, Abra_g_preds2, Abra_g_preds3, Abra_g_preds4)

# combined effect of bleaching severity and water temperature
Abra_g_preds5 <- ggpredict(Abra_best_mod, terms = c("log_2022 [all]", "temp_mean_z [-1, 0, 1]", "bl_cat1 [no bleaching]"), type = "fixed")
Abra_g_preds6 <- ggpredict(Abra_best_mod, terms = c("log_2022 [all]", "temp_mean_z [-1, 0, 1]", "bl_cat1 [minor bleaching]"), type = "fixed")
Abra_g_preds7 <- ggpredict(Abra_best_mod, terms = c("log_2022 [all]", "temp_mean_z [-1, 0, 1]", "bl_cat1 [major bleaching]"), type = "fixed")
Abra_g_preds8 <- ggpredict(Abra_best_mod, terms = c("log_2022 [all]", "temp_mean_z [-1, 0, 1]", "bl_cat1 [fully bleached]"), type = "fixed")
Abra_g_preds9 <- ggpredict(Abra_best_mod, terms = c("log_2022 [all]", "temp_mean_z [-1, 0, 1]", "bl_cat1 [recent partial mortality]"), type = "fixed")

Abra_g_predsb <- rbind(Abra_g_preds5, Abra_g_preds6, Abra_g_preds7, Abra_g_preds8, Abra_g_preds9)

# Replace these with actual mean and sd from grow_env_dat$temp_mean
temp_mean <- mean(grow_env_dat$temp_mean, na.rm = TRUE)
temp_sd <- sd(grow_env_dat$temp_mean, na.rm = TRUE)

# Translate z-scores to real temperatures
z_to_real <- function(z) round(temp_mean + z * temp_sd, 2)

Abra_g_predsb$temperature_c <- z_to_real(as.numeric(as.character(Abra_g_predsb$group)))
```

```{r}
#| label: Abra growth fig
#| echo: false
#| fig-width: 12
#| fig-height: 8
#| warning: false
#| message: false

colours_sev <- c("#F0E442",  "#E69F00", "#D55E00", "red", "black")
colours_temp <- c("#714674", "#FF99FF", "#FAAE7B")

# effect of bleaching severity
ggplot() +
  geom_abline(intercept = 0, 
              slope = 1, 
              linetype = "dashed", 
              color = "grey50", 
              size = 1) +
  geom_line(data = Abra_g_predsa, 
            aes(x = x, y = predicted, color = group), 
            size = 1) +
  geom_ribbon(data = Abra_g_predsa,
              aes(x = x,
                  ymin = conf.low,
                  ymax = conf.high,
                  fill = group),
              alpha = 0.2,
              inherit.aes = FALSE,
              show.legend = FALSE) +
  labs(x = "Colony size in 2022 (log)",
       y = "Colony size in 2023 (log)") +
  scale_color_manual(values = colours_sev) +
  scale_fill_manual(values = colours_sev) +
  scale_x_continuous(breaks = seq(2, 12, by = 2),
                     limits = c(2, 12)) +
  scale_y_continuous(breaks = seq(2, 12, by = 2),
                     limits = c(2, 12)) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12))

# Effect of water temperature and bleaching severity
ggplot() +
  geom_abline(intercept = 0, 
              slope = 1, 
              linetype = "dashed", 
              color = "grey50", 
              size = 1) +
  geom_line(data = Abra_g_predsb, 
            aes(x = x, y = predicted, color = factor(temperature_c)), 
            size = 1) +
  geom_ribbon(data = Abra_g_predsb,
              aes(x = x,
                  ymin = conf.low,
                  ymax = conf.high,
                  fill = factor(temperature_c)),
              alpha = 0.2,
              inherit.aes = FALSE,
              show.legend = FALSE) +
  facet_wrap(~facet) +
  labs(x = "",
       y = "",
       color = "") +
  scale_color_manual(values = colours_temp) +
  scale_fill_manual(values = colours_temp) +
  scale_x_continuous(breaks = seq(2, 12, by = 2),
                     limits = c(2, 12)) +
  scale_y_continuous(breaks = seq(2, 12, by = 2),
                     limits = c(2, 13)) +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12))
```

```{r}
#| label: Abra growth zero crossings
#| echo: false
Abra_g_predsb$diff <- Abra_g_predsb$predicted - Abra_g_predsb$x
Abra_zero_crossings <- Abra_g_predsb |>
  group_by(facet, group) |>
  arrange(abs(diff)) |>  
  slice(1)                
```

#### 2.2.3.b. Figure 5: Acropora corymbose

```{r}
#| label: acor summary
#| echo: false
summary(Acor_best_mod)
```

The best-fit model for *A. corymbose* included the interactions between initial size and bleaching category. Only the categories 'major bleaching' and 'recent partial mortality' were significantly different from 'no bleaching' category. The interaction between initial size and water velocity was also significant (negative effect).

```{r}
#| label: Acor preds
#| echo: false
#| warning: false
#| message: false

# Predicted effects of size across bleaching categories
Acor_g_preds0 <- ggpredict(Acor_best_mod, terms = c("log_2022 [all]", "bl_cat1 [no bleaching]"), type = "fixed")
Acor_g_preds1 <- ggpredict(Acor_best_mod, terms = c("log_2022 [all]", "bl_cat1 [minor bleaching]"), type = "fixed")
Acor_g_preds2 <- ggpredict(Acor_best_mod, terms = c("log_2022 [all]", "bl_cat1 [major bleaching]"), type = "fixed")
Acor_g_preds3 <- ggpredict(Acor_best_mod, terms = c("log_2022 [all]", "bl_cat1 [fully bleached]"), type = "fixed")
Acor_g_preds4 <- ggpredict(Acor_best_mod, terms = c("log_2022 [all]", "bl_cat1 [recent partial mortality]"), type = "fixed")

Acor_g_predsa <- rbind(Acor_g_preds0, Acor_g_preds1, Acor_g_preds2, Acor_g_preds3, Acor_g_preds4)

# Use z-values of -1, 0, 1 to represent low, average, high velocity
Acor_g_preds_ubed <- ggpredict(
  Acor_best_mod,
  terms = c("log_2022 [all]", "ubedmean_z [-1, 0, 1]", "bl_cat1 [no bleaching]"), 
  type = "fixed"
)
```

```{r}
#| label: Acor growth velocity fig
#| echo: false
#| warning: false
#| message: false

colours_velo <- c("#5EA8FF", "#2066CC", "#0B3D91")

# Convert to numeric and label
Acor_g_preds_ubed$ubed_velocity <- factor(Acor_g_preds_ubed$group, 
                                          labels = c("Low", "Mean", "High"))

ggplot(Acor_g_preds_ubed, aes(x = x, y = predicted, color = ubed_velocity)) +
  geom_line(size = 1) +
  geom_abline(intercept = 0, 
              slope = 1, 
              linetype = "dashed", 
              color = "grey50", 
              size = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = ubed_velocity), alpha = 0.2, color = NA) +
  labs(
    x = "log(Size at t₀)",
    y = "Predicted log(Size at t₁)"
    # color = "Mean velocity (m/s)",
    # fill = "Mean velocity (m/s)",
  ) +
  scale_color_manual(values = colours_velo) +
  scale_fill_manual(values = colours_velo) +
  scale_x_continuous(breaks = seq(2, 10, by = 2), limits = c(2, 9)) +
  scale_y_continuous(breaks = seq(2, 10, by = 2), limits = c(2, 9)) +
  theme_bw() +
  theme(
     legend.position = "none",
    # legend.justification = c("left", "top"),
    # #legend.background = element_rect(fill = "white", color = "black"),
    # legend.title = element_text(size = 12),
    # legend.text = element_text(size = 10),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12))
```

```{r}
#| label: Acor growth bl fig
#| echo: false
#| warning: false
#| message: false

colours_sev <- c("#F0E442",  "#E69F00", "#D55E00", "red", "black")

ggplot() +
  geom_abline(intercept = 0, 
              slope = 1, 
              linetype = "dashed", 
              color = "grey50", 
              size = 1) +
  geom_line(data = Acor_g_predsa, 
            aes(x = x, y = predicted, color = group), 
            size = 1) +
  geom_ribbon(data = Acor_g_predsa,
              aes(x = x,
                  ymin = conf.low,
                  ymax = conf.high,
                  fill = group),
              alpha = 0.2,
              inherit.aes = FALSE,
              show.legend = FALSE) +
  labs(x = "Colony size in 2022 (log)",
       y = "Colony size in 2023 (log)") +
  scale_color_manual(values = colours_sev) +
  scale_fill_manual(values = colours_sev) +
  scale_x_continuous(breaks = seq(2, 9, by = 2),
                     limits = c(2, 9)) +
  scale_y_continuous(breaks = seq(2, 9, by = 2),
                     limits = c(2, 9)) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12))
```

#### 2.2.3.c. Figure 6: Pocillopora damicornis

```{r}
#| label: pdam summary
#| echo: false
summary(Pdam_best_mod)
```

The best-fit model for *P. damicornis* included initial size interacting bleaching severity. Only the category 'recent partial mortality' was significantly different from 'no bleaching'.

```{r}
#| label: Pdam preds
#| echo: false
#| warning: false
#| message: false

Pdam_g_preds2a <- ggpredict(Pdam_best_mod, terms = c("log_2022 [all]", "bl_cat1 [no bleaching]"), type = "fixed")
Pdam_g_preds2b <- ggpredict(Pdam_best_mod, terms = c("log_2022 [all]", "bl_cat1 [minor bleaching]"), type = "fixed")
Pdam_g_preds2c <- ggpredict(Pdam_best_mod, terms = c("log_2022 [all]", "bl_cat1 [major bleaching]"), type = "fixed")
Pdam_g_preds2d <- ggpredict(Pdam_best_mod, terms = c("log_2022 [all]", "bl_cat1 [fully bleached]"), type = "fixed")
Pdam_g_preds2e <- ggpredict(Pdam_best_mod, terms = c("log_2022 [all]", "bl_cat1 [recent partial mortality]"), type = "fixed")

Pdam_g_preds2 <- rbind(Pdam_g_preds2a, Pdam_g_preds2b, Pdam_g_preds2c, Pdam_g_preds2d, Pdam_g_preds2e)
```

```{r}
#| label: Pdam growth fig
#| echo: false
#| fig-width: 12
#| fig-height: 8
#| warning: false
#| message: false

ggplot() +
  geom_abline(intercept = 0, 
              slope = 1, 
              linetype = "dashed", 
              color = "grey50", 
              size = 1) +
  geom_line(data = Pdam_g_preds2, 
            aes(x = x, y = predicted, color = group), 
            size = 1) +
  geom_ribbon(data = Pdam_g_preds2,
              aes(x = x,
                  ymin = conf.low,
                  ymax = conf.high,
                  fill = group),
              alpha = 0.2,
              inherit.aes = FALSE,
              show.legend = FALSE) +
  labs(x = "Colony size in 2022 (log)",
       y = "Colony size in 2023 (log)",
       color = "") +
  scale_color_manual(values = colours_sev) +
  scale_fill_manual(values = colours_sev) +
  scale_x_continuous(breaks = seq(2, 9, by = 2), limits = c(2, 9)) +
  scale_y_continuous(breaks = seq(2, 9, by = 2), limits = c(2, 9)) +
  theme_bw() +
  theme(
    # legend.position = c(0.1, 0.75),
    legend.position = "none",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12))

```

The best-fit model for *P. damicornis* was the null model including only the interaction between initial size and bleaching severity. Only the category "recent partial mortality" was significant but note there was only 4 colonies in this category.

#### 2.2.3.d. Figure 7: Pocillopora verrucosa

```{r}
#| label: pver summary
#| echo: false
summary(Pver_best_mod)
```

The best-fit model for *P. verrucosa* included mean water temperature interacting with initial size and bleaching severity. The three-way interaction between initial size, mean temp and bleaching severity categories 'fully bleached' and 'recent partial mortality' were significant, however there were only 2 out of 207 colonies in the 'recent partial mortality' category so we believe these results are not reliable.

```{r}
#| label: Pver preds
#| echo: false
#| warning: false
#| message: false
# effect of bleaching severity
Pver_g_preds0 <- ggpredict(Pver_best_mod, terms = c("log_2022 [all]", "bl_cat1 [no bleaching]"), type = "fixed")
Pver_g_preds1 <- ggpredict(Pver_best_mod, terms = c("log_2022 [all]", "bl_cat1 [minor bleaching]"), type = "fixed")
Pver_g_preds3 <- ggpredict(Pver_best_mod, terms = c("log_2022 [all]", "bl_cat1 [fully bleached]"), type = "fixed") #excludes recent partial mortality category as there are only 2 colonies in it and predictions are unreliable

Pver_g_predsa <- rbind(Pver_g_preds0, Pver_g_preds1, Pver_g_preds3)

# combined effect of bleaching severity and water temperature
Pver_g_preds5 <- ggpredict(Pver_best_mod, terms = c("log_2022 [all]", "temp_mean_z [-1, 0, 1]", "bl_cat1 [no bleaching]"), type = "fixed")
Pver_g_preds6 <- ggpredict(Pver_best_mod, terms = c("log_2022 [all]", "temp_mean_z [-1, 0, 1]", "bl_cat1 [minor bleaching]"), type = "fixed")
Pver_g_preds8 <- ggpredict(Pver_best_mod, terms = c("log_2022 [all]", "temp_mean_z [-1, 0, 1]", "bl_cat1 [fully bleached]"), type = "fixed")

Pver_g_predsb <- rbind(Pver_g_preds5, Pver_g_preds6, Pver_g_preds8)

# Replace these with actual mean and sd from grow_env_dat$temp_mean
temp_mean <- mean(grow_env_dat$temp_mean, na.rm = TRUE)
temp_sd <- sd(grow_env_dat$temp_mean, na.rm = TRUE)

# Translate z-scores to real temperatures
z_to_real <- function(z) round(temp_mean + z * temp_sd, 2)

Pver_g_predsb$temperature_c <- z_to_real(as.numeric(as.character(Pver_g_predsb$group)))
```

```{r}
#| label: Pver growth fig
#| echo: false
#| fig-width: 12
#| fig-height: 10
#| warning: false
#| message: false

colours_sev2 <- c("#F0E442",  "#E69F00", "red")
colours_temp <- c("#714674", "#FF99FF", "#FAAE7B")

# effect of bleaching severity
ggplot() +
  geom_abline(intercept = 0, 
              slope = 1, 
              linetype = "dashed", 
              color = "grey50", 
              size = 1) +
  geom_line(data = Pver_g_predsa, 
            aes(x = x, y = predicted, color = group), 
            size = 1) +
  geom_ribbon(data = Pver_g_predsa,
              aes(x = x,
                  ymin = conf.low,
                  ymax = conf.high,
                  fill = group),
              alpha = 0.2,
              inherit.aes = FALSE,
              show.legend = FALSE) +
  labs(x = "Colony size in 2022 (log)",
       y = "Colony size in 2023 (log)") +
  scale_color_manual(values = colours_sev2) +
  scale_fill_manual(values = colours_sev2) +
  scale_x_continuous(breaks = seq(3, 8, by = 2),
                     limits = c(3, 8)) +
  scale_y_continuous(breaks = seq(3, 8, by = 2),
                     limits = c(3, 8)) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12))
```

```{r}
#| label: Pver growth zero crossings
#| echo: false
Pver_g_predsa$diff <- Pver_g_predsa$predicted - Pver_g_predsa$x
zero_crossings <- Pver_g_predsa |>
  group_by(facet, group) |>
  arrange(abs(diff)) |>  
  slice(1)                
```

#### 2.2.3.e. Figure 8: Seriatopora hystrix

```{r}
#| label: shys summary
#| echo: false
summary(Shys_best_mod)
```

The best-fit model for *S. hystrix* included mean temperature interacting with initial size and bleaching severity. The three-way interaction between initial size, mean temperature and bleaching severity category was significant for the 'major bleaching' and 'fully bleached' categories.

```{r}
#| label: Shys preds
#| echo: false
#| warning: false
#| message: false
library(ggeffects)

# Generate predictions across size and temperature for each bleaching level
Shys_g_preds0 <- ggpredict(Shys_best_mod, terms = c("log_2022 [all]", "temp_mean_z [-1,0,1]", "bl_cat1 [no bleaching]"), type = "fixed")
Shys_g_preds1 <- ggpredict(Shys_best_mod, terms = c("log_2022 [all]", "temp_mean_z [-1,0,1]", "bl_cat1 [minor bleaching]"), type = "fixed")
Shys_g_preds2 <- ggpredict(Shys_best_mod, terms = c("log_2022 [all]", "temp_mean_z [-1,0,1]", "bl_cat1 [major bleaching]"), type = "fixed")
Shys_g_preds3 <- ggpredict(Shys_best_mod, terms = c("log_2022 [all]", "temp_mean_z [-1,0,1]", "bl_cat1 [fully bleached]"), type = "fixed")
Shys_g_preds4 <- ggpredict(Shys_best_mod, terms = c("log_2022 [all]", "temp_mean_z [-1,0,1]", "bl_cat1 [recent partial mortality]"), type = "fixed")

Shys_g_preds <- dplyr::bind_rows(
  Shys_g_preds0,
  Shys_g_preds1,
  Shys_g_preds2,
  Shys_g_preds3,
  Shys_g_preds4
)
```

```{r}
#| label: Pver growth fig
#| echo: false
#| fig-width: 12
#| fig-height: 10
#| warning: false
#| message: false

colours_temp <- c("#714674", "#FF99FF", "#FAAE7B")

ggplot() +
  geom_abline(intercept = 0, 
              slope = 1, 
              linetype = "dashed", 
              color = "grey50", 
              size = 1) +
  geom_line(data = Shys_g_preds, 
            aes(x = x, y = predicted, color = group), 
            size = 1) +
  geom_ribbon(data = Shys_g_preds,
              aes(x = x,
                  ymin = conf.low,
                  ymax = conf.high,
                  fill = group),
              alpha = 0.2,
              inherit.aes = FALSE,
              show.legend = FALSE) +
  facet_wrap(~facet, scales = "free") +
  labs(x = "Colony size in 2022 (log)",
       y = "",
       color = "mean temp (C)") +
  scale_color_manual(values = colours_temp) +
  scale_fill_manual(values = colours_temp) +
    scale_x_continuous(breaks = seq(2, 8, by = 2), limits = c(2, 8)) +
  scale_y_continuous(breaks = seq(2, 8, by = 2), limits = c(1, 10)) +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12))

Shys_g_plotA + Shys_g_plotC+ plot_layout(nrow = 2)
```

```{r}
#| label: Shys growth zero crossings
#| echo: false
Shys_g_preds$diff <- Shys_g_preds$predicted - Shys_g_preds$x
shys_zero_crossings <- Shys_g_preds |>
  group_by(facet, group) |>
  arrange(abs(diff)) |>  
  slice(1)                
```

# 3. Supplementary analyses


```{r}
#| label: ReadData survival
#| output: false
#| echo: false
#| eval: true
#| warning: false
#| message: false
surv2 <- read_csv("C:/Users/jc815475/OneDrive - James Cook University/PhD data/3_Coral growth/Data/CSV/survival_transition2.csv", trim_ws=TRUE)
```

```{r}
#| label: survival prep
#| output: false
#| echo: false
#| eval: true
#| warning: false
#| message: false
surv_2 <- surv2 |> 
  mutate(Reef = factor(Reef),
         Site = factor(Site),
         Plot = factor(Plot),
         Colony = factor(Colony),
         Habitat = factor(Habitat),
         Depth = factor(Depth),
         Cross_shelf = factor(Cross_shelf),
         Taxa = factor(Taxa))|> 
  filter(Taxa %in% c("Abra", "Acor", "Pdam", "Pver", "Shys"))

surv_2 <- surv_2 |> rename(Survival = "Survival_2",
                           log_size = "log_2022")

surv_2_sub <- surv_2 |> 
  filter(!Bleaching_scores == "6")

surv_2_sub <- surv_2_sub |> 
  mutate(Bleaching_scores = factor(Bleaching_scores),
         bl_cat1 = fct_recode(Bleaching_scores,
                              `no bleaching` = "1",
                              `minor bleaching` = "2",
                              `major bleaching` = "3",
                              `fully bleached` = "4",
                              `recent partial mortality` = "5"),
         bl_cat2 = fct_recode(Bleaching_scores,
                              `no bleaching` = "1",
                              `partial bleaching` = "2",
                              `partial bleaching` = "3",
                              `fully bleached` = "4",
                              `recent partial mortality` = "5"))
```

For these taxa we decided to filter out colonies smaller than 5 cm in diameter (n = 2163 - 2126 = 37 colonies)

```{r}
#| label: survival prep2
#| output: false
#| echo: false
#| eval: true
#| warning: false
#| message: false

#Filter out colonies that are less than 5cm in diameter
# Area = πr^2
# Where π (pi) is approximately 3.14159, and r is the radius of the circle.
# So if we assume colonies are circle, and we want to filter out the colonies with D < 5 cm, the radius (r) is half of the diameter, so r = 2.5 cm / 2 = 5 cm.
# Calculate the area of a circle R = 2.5 cm:
# Area = π * (2.5 cm)^2
# Area ≈ 78.54 square centimeters
# So, the area of a circle with a diameter of 5 centimeters is approximately 19.6 square centimeters

surv_2_sub <- surv_2_sub |> 
  filter(Area_cm_2022 >= 19.6)

surv_2_sub |> 
  group_by(Taxa) |> 
  summarise(total_count = n()) #lowest sample size is Pdam with 208 colonies
```

## 3.1. Base model for Survival model

### 3.1.a. Acropora staghorn

```{r}
#| label: model1.1.a selection 
Abra_s0 <- glmmTMB(Survival ~ log_size : bl_cat1 +
                (1|Site),
              data = surv_2_sub[surv_2_sub$Taxa == "Abra",],
              family = "binomial")

Abra_s1 <- glmmTMB(Survival ~ bl_cat1 + log_size +
                (1|Site),
              data = surv_2_sub[surv_2_sub$Taxa == "Abra",],
              family = "binomial")

Abra_s2 <- glmmTMB(Survival ~ bl_cat1 +
                (1|Site),
              data = surv_2_sub[surv_2_sub$Taxa == "Abra",],
              family = "binomial")
```

```{r}
#| label: model1.1.a AIC 
AIC(Abra_s0, Abra_s1, Abra_s2)
```

For *A. branching*, the best-fit model includes the interaction between initial size and bleaching severity.

### 3.1.b. Acropora corymbose

```{r}
#| label: model1.1.b selection 
Acor_s0 <- glmmTMB(Survival ~ bl_cat1 : log_size +
                (1|Site),
              data = surv_2_sub[surv_2_sub$Taxa == "Acor",],
              family = "binomial")

Acor_s1 <- glmmTMB(Survival ~ bl_cat1 + log_size +
                (1|Site),
              data = surv_2_sub[surv_2_sub$Taxa == "Acor",],
              family = "binomial")

Acor_s2 <- glmmTMB(Survival ~ bl_cat1 +
                (1|Site),
              data = surv_2_sub[surv_2_sub$Taxa == "Acor",],
              family = "binomial")
```

```{r}
AIC(Acor_s0, Acor_s1, Acor_s2)
```

For *A. corymbose*, the best-fit model includes initial size and bleaching severity but not their interaction.

### 3.1.c. Pocillopora damicornis

```{r}
#| label: model1.1.c selection
Pdam_s0 <- glmmTMB(Survival ~ bl_cat1 : log_size +
                (1|Site),
              data = surv_2_sub[surv_2_sub$Taxa == "Pdam",],
              family = "binomial")

Pdam_s1 <- glmmTMB(Survival ~ bl_cat1 + log_size +
                (1|Site),
              data = surv_2_sub[surv_2_sub$Taxa == "Pdam",],
              family = "binomial")

Pdam_s2 <- glmmTMB(Survival ~ bl_cat1 +
                (1|Site),
              data = surv_2_sub[surv_2_sub$Taxa == "Pdam",],
              family = "binomial")
```

```{r}
AIC(Pdam_s0, Pdam_s1, Pdam_s2)
```

For *P. damicornis*, the best-fit model includes initial size and bleaching severity but not their interaction.

### 3.1.d. Pocillopora verrucosa

```{r}
#| label: model1.1.d selection
Pver_s0 <- glmmTMB(Survival ~ bl_cat1 : log_size +
                (1|Site),
              data = surv_2_sub[surv_2_sub$Taxa == "Pver",],
              family = "binomial")
Pver_s1 <- glmmTMB(Survival ~ bl_cat1 + log_size +
                (1|Site),
              data = surv_2_sub[surv_2_sub$Taxa == "Pver",],
              family = "binomial")
Pver_s2 <- glmmTMB(Survival ~ bl_cat1 +
                (1|Site),
              data = surv_2_sub[surv_2_sub$Taxa == "Pver",],
              family = "binomial")
```

```{r}
AIC(Pver_s0, Pver_s1, Pver_s2)
```

For *P. verrucosa*, the best-fit model includes initial size and bleaching severity but not their interaction.

### 3.1.e. Seriatopora hystrix

```{r}
#| label: model1.1.e selection
Shys_s0 <- glmmTMB(Survival ~ bl_cat1 : log_size +
                (1|Site),
              data = surv_2_sub[surv_2_sub$Taxa == "Shys",],
              family = "binomial")
Shys_s1 <- glmmTMB(Survival ~ bl_cat1 + log_size +
                (1|Site),
              data = surv_2_sub[surv_2_sub$Taxa == "Shys",],
              family = "binomial")
Shys_s2 <- glmmTMB(Survival ~ bl_cat1 +
                (1|Site),
              data = surv_2_sub[surv_2_sub$Taxa == "Shys",],
              family = "binomial")
```

```{r}
AIC(Shys_s0, Shys_s1, Shys_s2)
```

For *S. hystrix*, the best-fit model includes initial size interacting with bleaching severity.

## 3.1.f. Summary figures

```{r}
#| label: model1.1.predictions
#| echo: false
Abra_s0_preds <- predict_response(Abra_s0, terms = c("log_size [all]", "bl_cat1"), margin = "marginalmeans")
Acor_s1_preds <- predict_response(Acor_s1, terms = "log_size [all]", margin = "marginalmeans")
Pdam_s1_preds <- predict_response(Pdam_s1, terms = c("log_size [all]"), margin = "marginalmeans")
Pver_s1_preds <- predict_response(Pver_s1, terms = c("log_size [all]"), margin = "marginalmeans")
Shys_s0_preds <- predict_response(Shys_s0, terms = c("log_size [all]", "bl_cat1"), margin = "marginalmeans")
```

```{r}
#| label: model1.1.predictions cont.
#| echo: false
Abra_s0_preds$Taxa = "Abra"
Pdam_s1_preds$Taxa = "Pdam"
Pver_s1_preds$Taxa = "Pver"
Acor_s1_preds$Taxa = "Acor"
Shys_s0_preds$Taxa = "Shys"

group_int <- rbind(Abra_s0_preds, Shys_s0_preds)
group_no_int <- rbind(Acor_s1_preds, Pdam_s1_preds, Pver_s1_preds)
```

```{r}
#| label: model1.1. pred for bl_cat
#| echo: false
#| results: false

surv_bl_preds1 <- predict_response(Acor_s1, terms = "bl_cat1 [all]", margin = "marginalmeans")
surv_bl_preds1$Taxa = "Acor"
surv_bl_preds2 <- predict_response(Pdam_s1, terms = "bl_cat1 [all]", margin = "marginalmeans")
surv_bl_preds2$Taxa = "Pdam"
surv_bl_preds3 <- predict_response(Pver_s1, terms = "bl_cat1 [all]", margin = "marginalmeans")
surv_bl_preds3$Taxa = "Pver"
surv_bl_preds <- rbind(surv_bl_preds1, surv_bl_preds2, surv_bl_preds3)

surv_bl_preds |> 
  mutate(x = factor(x, levels = c("no bleaching", "minor bleaching", "major bleaching", "fully bleached", "recent partial mortality")))
surv_bl_preds |> mutate(Taxa = factor(Taxa))
```

```{r}
#| label: model1.1. figure
#| fig-width: 10
#| fig-height: 10

taxa_labels1 <- c("Abra" = "A. branching",
                 "Shys" = "S. hystrix")

taxa_labels2 <- c("Acor" = "A. corymbose",
                 "Pdam" = "P. damicornis",
                 "Pver" = "P. verrucosa")

colours_sev = c("#F0E442",  "#E69F00", "#D55E00", "red", "black")

surv_plotA <- ggplot(group_int, aes(x = x, y = predicted, color = group)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group), alpha = 0.2, color = NA) +
  facet_wrap(~Taxa, labeller =  as_labeller(taxa_labels1), ) +
  scale_fill_manual(values = colours_sev) +
  scale_color_manual(values = colours_sev) +
  labs(title = "A) ", 
       x = "Log Area 2022", 
       y = "Survival Probability", 
       color = "Bleaching Category", 
       fill = "Bleaching Category") +
  theme_bw() +
  xlab("initial size (log)") +
  theme(text = element_text(size = 12),
        panel.grid.major.y = element_line(linetype = "dashed"),
        strip.text = element_text(size = 12),
        legend.position = "none"
  )

surv_plotB <- ggplot(group_no_int, aes(x = x, y = predicted)) +
  geom_line(linewidth = 1.2, color = "blue") +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, color = NA) +
  facet_wrap(~Taxa, labeller =  as_labeller(taxa_labels2), ) +
  labs(title = "B) ", 
       x = "initial size (log)", 
       y = "") +
  theme_bw() +
  scale_y_continuous(breaks = seq(0, 1, by = 0.25)) +
  theme(text = element_text(size = 12),
        panel.grid.major.y = element_line(linetype = "dashed"),
        strip.text = element_text(size = 12)
  )

surv_plotC <- ggplot()+
  geom_errorbar(data = surv_bl_preds, aes(x = x, ymin = conf.low, ymax = conf.high, 
                                   group = Taxa), width = 0.1, linewidth = 0.8) +
  geom_point(data = surv_bl_preds, aes(x = x, y = predicted, fill = x, group = Taxa),
             size = 4, pch = 21) +
  scale_colour_manual(values = colours_sev,
                      name = "") +
  scale_fill_manual(values = colours_sev,
                    name = "") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.25)) +
  theme_bw() +
  facet_wrap(~ Taxa, labeller =  as_labeller(taxa_labels2), )+
  labs(title = "C)", 
       x = "", 
       y = "Survival probability") +
  theme(
    text = element_text(size = 12),
    axis.text.x = element_blank(),
    panel.grid.major.y = element_line(linetype = "dashed"),
    panel.grid.minor.y = element_blank(),
    strip.text = element_text(size = 12),
    legend.position = "bottom",
    legend.text = element_text(size = 12)
  )+
  scale_x_discrete(limits = levels(surv_bl_preds$x))

surv_plotA + surv_plotB + surv_plotC + plot_layout(ncol = 2)

```

## 3.2. Data exploration - raw data plotting

### 3.2.a. Mortality rates pre- and post-bleaching (raw data plotting)

```{r}
#| label: Barplot1
#| echo: false
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 10

surv1_summary <- surv_1 |>
  group_by(Taxa) |>
  summarise(
    dead_count = sum(Survival_1 == 0),
    alive_count = sum(Survival_1 == 1),
    total_count = n(),
    dead_proportion = dead_count / total_count,
    alive_proportion = alive_count / total_count,
    .groups = 'drop'
  ) |>
  arrange(desc(dead_proportion))


surv1_long <- surv1_summary |>
  pivot_longer(cols = c(dead_proportion, alive_proportion), 
               names_to = "status", 
               values_to = "proportion")


surv1_long <- surv1_long |>
  group_by(Taxa) |>
  mutate(cumulative = cumsum(proportion))

surv1_long <- surv1_long |> 
  mutate(Taxa = factor(Taxa, levels = c("Abra", "Acor", "Pdam", "Pver","Shys")))

mort1 <- ggplot(surv1_long, aes(x = Taxa, y = proportion, fill = status)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("dead_proportion" = "black", "alive_proportion" = "grey"), 
                    labels = c("dead_proportion" = "Dead", "alive_proportion" = "Alive")) +
  labs(x = "", y = "Proportion of colonies", fill = "Survival", title = "A - year 1 (no bleaching)") +
  geom_text(aes(label = scales::percent(proportion, accuracy = 0.1), 
                y = cumulative - proportion / 2), 
            color = "white") +
  theme_bw() +
    scale_x_discrete(limits = c("Abra", "Acor", "Pdam", "Pver","Shys"),
                   labels = c("Acropora branching",
                              "Acropora corymbose",
                              "Pocillopora damicornis",
                              "Pocillopora verrucosa",
                              "Seriatopora hystrix")) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 90, hjust = 1, size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_blank()
  )



surv2_summary <- surv_2 |>
  group_by(Taxa) |>
  summarise(
    dead_count = sum(Survival_2 == 0),
    alive_count = sum(Survival_2 == 1),
    total_count = n(),
    dead_proportion = dead_count / total_count,
    alive_proportion = alive_count / total_count,
    .groups = 'drop'
  ) |>
  arrange(desc(dead_proportion))


surv2_long <- surv2_summary |>
  pivot_longer(cols = c(dead_proportion, alive_proportion), 
               names_to = "status", 
               values_to = "proportion")


surv2_long <- surv2_long |>
  group_by(Taxa) |>
  mutate(cumulative = cumsum(proportion))

surv2_long <- surv2_long |> 
  mutate(Taxa = factor(Taxa, levels = c("Abra", "Acor", "Pdam", "Pver","Shys")))

mort2 <- ggplot(surv2_long, aes(x = Taxa, y = proportion, fill = status)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("dead_proportion" = "black", "alive_proportion" = "grey"), 
                    labels = c("dead_proportion" = "Dead", "alive_proportion" = "Alive")) +
  labs(x = "", y = "Proportion of colonies", fill = "Survival", title = "B - year 2 (bleaching)") +
  geom_text(aes(label = scales::percent(proportion, accuracy = 0.1), 
                y = cumulative - proportion / 2), 
            color = "white") +
  theme_bw() +
    scale_x_discrete(limits = c("Abra", "Acor", "Pdam", "Pver","Shys"),
                   labels = c("Acropora branching",
                              "Acropora corymbose",
                              "Pocillopora damicornis",
                              "Pocillopora verrucosa",
                              "Seriatopora hystrix")) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 90, hjust = 1, size = 12),
    legend.text = element_text(size = 12),
    legend.title = element_blank()
  )

mort1 + mort2 + plot_layout(ncol = 1)

```

Mortality increased after bleaching for all taxa.

### 3.2.b. Survival rates per bleaching category (raw data plotting)

```{r}
#| label: survHeatmap
#| echo: false
#| message: false
#| warning: false
#| fig-width: 14
#| fig-height: 8

surv_2_sub <- surv_2 |> 
  filter(!Bleaching_scores == "6")

taxa_labels <- c("Abra" = "Acropora branching",
                 "Acor" = "Acropora corymbose",
                 "Pdam" = "Pocillopora damicornis",
                 "Pver" = "Pocillopora verrucosa",
                 "Shys" = "Seriatopora hystrix")

bl_cat <- c("1" = "No bleaching",
            "2" = "Minor bleaching",
            "3" = "Major bleaching",
            "4" = "Fully bleached",
            "5" = "Recent partial mortality")

Taxa_sum2 <- surv_2_sub |> 
  group_by(Taxa, Bleaching_scores) |> 
  summarize(Survival_Rate2 = mean(Survival_2),
            Colony_count2 = n())

ggplot(Taxa_sum2, aes(x = factor(Bleaching_scores), y = Taxa, fill = Survival_Rate2)) +
  geom_tile() +
  geom_text(aes(label = Colony_count2), color = "white", size = 5) +
  scale_fill_gradient(low = "black", high = "grey", limits = c(min_surv, max_surv)) +
  scale_y_discrete(limits = c("Shys", "Pver","Pdam","Acor","Abra"),
                   labels = c("Seriatopora hystrix",
                              "Pocillopora verrucosa",
                              "Pocillopora damicornis",
                              "Acropora corymbose",
                              "Acropora branching")) +
  scale_x_discrete(labels = bl_cat) +
  labs(x = "", 
       y = "", 
       fill = "Survival Rate") +
  theme_classic()+
  theme(
    axis.text.x = element_text(size = 12, angle = 90),
    axis.title.y = element_text(size = 12),
    legend.text = element_text(size =12),
    legend.title = element_text(size = 12)
  )
```

### 3.2.c. Growth rates pre- and post-bleaching (all colonies)

```{r}
#| label: growfig1
#| echo: false
#| message: false
#| warning: false
#| fig-width: 14
#| fig-height: 16
gt1 <- grow1 |> ggplot(
  aes(x = log10_2021, y = log10_2022, group = Taxa)) +
  geom_point(alpha = 0.1, pch = 16) +
  geom_smooth(method = "lm", color = "deepskyblue")+
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(title = "A - Year 1 (pre-bleaching)",
       x = "Size in 2021 (log)",
       y = "Size in 2022 (log)") +
  # scale_x_continuous(limits = c(0, 12), breaks = seq(0, 12, by = 2)) +
  # scale_y_continuous(limits = c(0, 12), breaks = seq(0, 12, by = 2)) +
  facet_wrap(~Taxa, scales = "free", labeller =  as_labeller(taxa_labels), ) +
  theme_bw()

gt2 <- grow2 |> ggplot(
  aes(x = log10_2022, y = log10_2023, group = Taxa)) +
  geom_point(alpha = 0.1, pch = 16) +
  geom_smooth(method = "lm", color = "darkorange")+
  geom_abline(slope = 1, intercept = 0) +
  labs(title = "B - Year 2 (post-bleaching)",
       x = "Size in 2022 (log)",
       y = "Size in 2023 (log)") +
  # scale_x_continuous(limits = c(0, 12), breaks = seq(0, 12, by = 2)) +
  # scale_y_continuous(limits = c(0, 12), breaks = seq(0, 12, by = 2)) +
  facet_wrap(~Taxa, scales = "free", labeller =  as_labeller(taxa_labels), ) +
  theme_bw()

gt1 + gt2 + plot_layout(ncol = 1)
```

The dashed line represents the no-growth line (intercept 0, slope 1). All slopes are \< 1 meaning growth rates decreases with size. It looks like in general in year 2, the curves are closer to the no-growth line.

### 3.2.d. Growth rates pre- and post-bleaching (healthy colonies only)

```{r}
#| label: healthydataprep
#| echo: false
#| message: false
#| warning: false
#| results: hide
rawdata <- read_csv("C:/Users/jc815475/OneDrive - James Cook University/PhD data/3_Coral growth/Data/CSV/growth_surv_data.csv", trim_ws=TRUE)

dat1 <- rawdata |> 
  mutate(Taxa_2023 = factor(Class_2023),
         Taxa_2023 = fct_recode(Taxa_2023,
                                Atab = "Acyt",
                                Atab = "Ahya",
                                Abra = "Asta",
                                Adgt = "Agem",
                                Adgt = "Ahum",
                                Acor = "Amil",
                                Acor = "Aten",
                                Gspp = "Gedw",
                                Gspp = "Gret"),
         Taxa_2022 = factor(Class_2022),
         Taxa_2022 = fct_recode(Taxa_2022,
                                Atab = "Acyt",
                                Atab = "Ahya",
                                Abra = "Asta",
                                Adgt = "Agem",
                                Adgt = "Ahum",
                                Acor = "Amil",
                                Acor = "Aten",
                                Gspp = "Gedw",
                                Gspp = "Gret"),
         Taxa_2021 = factor(Class_2021),
         Taxa_2021 = fct_recode(Taxa_2021,
                                Atab = "Acyt",
                                Atab = "Ahya",
                                Abra = "Asta",
                                Adgt = "Agem",
                                Adgt = "Ahum",
                                Acor = "Amil",
                                Acor = "Aten",
                                Gspp = "Gedw",
                                Gspp = "Gret"))

### Subset data for each transition
healthy <- dat1 |> 
  filter(to_use_0 == "yes" & to_use_growth_1 == "yes" & to_use_growth_2 == "yes")

### Only retain taxa of interest
healthy <- healthy |> 
  filter(Taxa_2023 %in% c("Abra", "Acor", "Pdam", "Pver", "Shys")) |> 
  droplevels()

healthy <- healthy |> 
  filter(Taxa_2022 %in% c("Abra", "Acor", "Pdam", "Pver", "Shys")) |> 
  droplevels()

healthy <- healthy |> 
  filter(Taxa_2021 %in% c("Abra", "Acor", "Pdam", "Pver", "Shys")) |> 
  droplevels()

### only retain rows where Taxa match throughout the years
healthy <- healthy[healthy$Taxa_2021 == healthy$Taxa_2022 & healthy$Taxa_2022 == healthy$Taxa_2023, ]

### Check taxa levels for each year
levels(healthy$Taxa_2021)
levels(healthy$Taxa_2022)
levels(healthy$Taxa_2023)

### only retain colonies that stayed healthy in Year 2
healthy <- healthy |> 
  mutate(Bleaching_scores = factor(Bleaching_scores)) |> 
  filter(Bleaching_scores == "1")

## Setup hierarchical design
healthy <- healthy |> 
  mutate(Plot = factor(Plot_ID),
         Zone = factor(str_sub(Plot, 9,9)),
         Reef = factor(str_sub(Plot, 1,4)),
         Site = factor(str_sub(Plot, 1,9)), #unique combination of Reef/Site/Zone
         Site_type = factor(str_sub(Plot, 6, 7)),
         Colony_ID = row_number(),
         Colony = factor(paste(Plot, Colony_ID, sep = "_")),
         Habitat = factor(paste(Site_type, Zone, sep = "_")),
         Reef = fct_recode(Reef,
                           PA = "PAPE",
                           PA = "PAOR"),
         Depth = fct_recode(Zone,
                            Deep = "D",
                            Shallow = "S"),
         Area_cm_2021 = Area_2021*10000, #change area in m2 to cm2
         Area_cm_2022 = Area_2022*10000,
         Area_cm_2023 = Area_2023*10000,
         log_2021= log(Area_cm_2021),
         log_2022= log(Area_cm_2022),
         log_2023= log(Area_cm_2023),
         log10_2021=log10(Area_cm_2021),
         log10_2022 =log10(Area_cm_2022),
         log10_2023 =log10(Area_cm_2023),
         Taxa = factor(Taxa_2021),
         Cross_shelf = fct_recode(Reef,
                                  offshore = "OCCH",
                                  offshore = "OCDA",
                                  offshore = "OCLB",
                                  inshore = "PA"))

healthy <- healthy |> 
  select(Reef, Site, Plot, Colony, Habitat, Depth, Cross_shelf, Taxa, Area_cm_2021, Area_cm_2022, Area_cm_2023, log_2021, log_2022, log_2023, log10_2021, log10_2022, log10_2023, Note_2021, Note_2022, Note_2023)

```

```{r}
#| label: healthyrawfig
#| echo: false
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 10
# Pivot longer to reshape the dataset for modeling
healthy_long <- healthy |> 
  pivot_longer(cols = matches(".*_[0-9]{4}"),  
               names_to = c(".value", "Year"), 
               names_pattern = "(.+)_(.+)")    

healthy_long <- healthy_long |> 
  mutate(Year = factor(Year),
         Taxa = factor(Taxa))


# Plot raw data
data_2021_2022 <- healthy |>
  select(Taxa, log10_2021, log10_2022) |>
  rename(log_previous = log10_2021, log_size = log10_2022) |>
  mutate(Year_transition = "2021-2022")

data_2022_2023 <- healthy |>
  select(Taxa, log10_2022, log10_2023) |>
  rename(log_previous = log10_2022, log_size = log10_2023) |>
  mutate(Year_transition = "2022-2023")

plot_data <- bind_rows(data_2021_2022, data_2022_2023)

ggplot(plot_data, aes(x = log_previous, y = log_size, color = Year_transition)) +
  geom_point(alpha = 0.1, size = 2) +
  geom_smooth(method = "lm", se = FALSE) +  
  facet_wrap(~ Taxa, scales = "free", labeller =  as_labeller(taxa_labels)) +  
  labs(x = "log(size at t previous)", y = "log(size at t)") +
  theme_bw() +
  theme(
    legend.title = element_blank(),
    legend.position = "bottom"
  ) +
  scale_color_manual(values = c("2021-2022" = "cadetblue", "2022-2023" = "#D55E00"))
```

### 3.2.f. Growth per taxa and bleaching category

```{r}
#| label: growfig2
#| echo: false
#| message: false
#| warning: false
#| fig-width: 12
#| fig-height: 10
grow2 <- grow2 |> 
  mutate(bl_cat1 = case_when(
    is.na(Bleaching_scores)  ~ "no bleaching",
    Bleaching_scores == 1   ~ "no bleaching",
    Bleaching_scores == 2   ~ "minor bleaching",
    Bleaching_scores == 3   ~ "major bleaching",
    Bleaching_scores == 4   ~ "fully bleached",
    Bleaching_scores == 5   ~ "recent partial mortality"
  ))

grow2 <- grow2 |> 
  mutate(bl_cat1 = factor(bl_cat1, levels = c("no bleaching", "minor bleaching", "major bleaching", "fully bleached", "recent partial mortality")))

colours_bl = c("#F0E442","#E69F00", "#D55E00" ,"#990000", "black")

grow2 |> ggplot(
  aes(x = log10_2022, y = log10_2023, group = Taxa, colour = bl_cat1)) +
  geom_point(position = position_jitter(width = 0.4, height = 0.4), alpha = 0.2, pch = 16) +
  scale_color_manual(values = colours_bl) +
  labs(x = "Size in 2022 (log)",
       y = "Size in 2023 (log)") +
  # scale_x_continuous(limits = c(0, 12), breaks = seq(0, 12, by = 2)) +
  # scale_y_continuous(limits = c(0, 12), breaks = seq(0, 12, by = 2)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_smooth(aes(group = bl_cat1, colour = bl_cat1), method = "lm", se = FALSE, size = 1.2)+
  facet_wrap(~Taxa, scales = "free", labeller =  as_labeller(taxa_labels), ) +
  theme_bw() +
  theme(
    legend.title = element_blank()
  )
```
